---
title: "Don't Repeat Yourself (Unless You're a Politician)"
subtitle: "Functions and Iteration"
author: "Rod Albuyeh"
date: today
format: pdf
execute:
  echo: true
  eval: true
  warning: false
  error: false
  cache: true
  tidy: true
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 8, fig.height = 6)
```

## Learning Objectives

By the end of this lab, you will be able to:

- **Design** custom functions to solve policy-relevant problems
- **Apply** iteration techniques (for loops and functional programming) to automate repetitive tasks
- **Evaluate** trade-offs between different approaches to iteration
- **Implement** efficient workflows for processing multiple datasets or scenarios

## Setup

```{r libraries}
library(tidyverse)
library(scales)
library(here)

# Set a seed for reproducibility
set.seed(42)
```

# Part I: The Power of Functions in Policy Work

## Why Functions Matter for Data Science

In data science work, we often need to:

- Apply the same analysis to multiple datasets or groups
- Test different parameters or scenarios systematically
- Ensure consistent calculations across different parts of our analysis
- Make our work reproducible and easy to modify
- Reduce errors by eliminating copy-paste coding

Functions help us achieve all of these goals by encapsulating logic that we use repeatedly.

## The Function Development Workflow

**You don't necessarily have to startby writing a function!** Instead, follow this practical approach:

1. **Start with working code** for a specific case
2. **Identify what varies** between different uses
3. **Turn the varying parts into arguments**
4. **Wrap it in a function**

### Example: From Concrete to Abstract

Let's say we want to analyze unemployment data for different states:

```{r function-development, eval=FALSE}
# Step 1: Start with concrete code that works
california_data <- unemployment_data %>%
  filter(state == "CA") %>%
  summarise(
    avg_rate = mean(unemployment_rate),
    max_rate = max(unemployment_rate),
    trend = cor(year, unemployment_rate)
  )

# Step 2: Copy and modify for another state - notice what changes
texas_data <- unemployment_data %>%
  filter(state == "TX") %>%  # This changes!
  summarise(
    avg_rate = mean(unemployment_rate),  # This stays the same
    max_rate = max(unemployment_rate),   # This stays the same
    trend = cor(year, unemployment_rate) # This stays the same
  )

# Step 3: The varying part (state) becomes the function argument
analyze_state_unemployment <- function(data, state_code) {
  data %>%
    filter(state == state_code) %>%
    summarise(
      avg_rate = mean(unemployment_rate),
      max_rate = max(unemployment_rate),
      trend = cor(year, unemployment_rate)
    )
}

# Step 4: Now we can easily analyze any state
california_data <- analyze_state_unemployment(unemployment_data, "CA")
texas_data <- analyze_state_unemployment(unemployment_data, "TX")
```

**Key insight**: Start with copy-paste code, then look for patterns. The parts that change become your function arguments.

## Anatomy of a Function

```{r function-anatomy, eval=FALSE}
function_name <- function(input1, input2 = default_value) {
  # Do something with the inputs
  result <- some_calculation(input1, input2)
  
  # Return the result
  return(result)
}
```

## Example: Policy Impact Calculator

Let's create a function to calculate the impact of a minimum wage increase:

```{r wage-impact-function}
calculate_wage_impact <- function(current_wage, new_wage, affected_workers) {
  # Calculate the wage increase
  wage_increase <- new_wage - current_wage
  
  # Calculate total additional wages per year (assuming full-time work)
  annual_hours <- 40 * 52  # 40 hours/week * 52 weeks/year
  total_impact <- wage_increase * affected_workers * annual_hours
  
  # Create a summary
  impact_summary <- list(
    wage_increase = wage_increase,
    affected_workers = affected_workers,
    total_annual_impact = total_impact,
    impact_per_worker = wage_increase * annual_hours
  )
  
  return(impact_summary)
}

# Test the function
impact_result <- calculate_wage_impact(
  current_wage = 15.00,
  new_wage = 18.00,
  affected_workers = 50000
)

# Display results nicely
cat("Minimum Wage Impact Analysis\n")
cat("============================\n")
cat("Wage increase per hour: $", impact_result$wage_increase, "\n")
cat("Workers affected: ", scales::comma(impact_result$affected_workers), "\n")
cat("Additional income per worker annually: $", scales::comma(impact_result$impact_per_worker), "\n")
cat("Total additional wages annually: $", scales::comma(impact_result$total_annual_impact), "\n")
```

## Your Turn: Housing Affordability Function

**Task 1**: Inspect this function called `assess_housing_affordability()` that:

- Takes median income and median rent as inputs
- Calculates the percentage of income spent on rent (assume rent × 12 for annual)
- Returns "Affordable" if < 30%, "Moderately Unaffordable" if 30-50%, "Severely Unaffordable" if > 50%
- Also returns the exact percentage

```{r housing-function}

assess_housing_affordability <- function(median_income, median_rent) {
  # Calculate annual rent
  annual_rent <- median_rent * 12
  
  # Calculate percentage of income spent on rent
  rent_percentage <- (annual_rent / median_income) * 100
  
  # Determine affordability category
  if (rent_percentage < 30) {
    category <- "Affordable"
  } else if (rent_percentage <= 50) {
    category <- "Moderately Unaffordable"
  } else {
    category <- "Severely Unaffordable"
  }
  
  # Return results
  return(list(
    rent_percentage = rent_percentage,
    category = category,
    annual_rent = annual_rent,
    median_income = median_income
  ))
}

test_result <- assess_housing_affordability(median_income = 60000, median_rent = 2000)
cat("Housing affordability: ", test_result$category, "\n")
cat("Rent as % of income: ", round(test_result$rent_percentage, 1), "%\n")
```

# Part II: Iteration - Doing Things Repeatedly

## The For Loop: When You Need Sequential Control

For loops are the right choice when you need to:
- **Build results incrementally** where each iteration depends on the previous one
- **Access both the value and its position** in the data
- **Handle complex conditional logic** that varies by iteration
- **Modify objects in place** (though this should be used sparingly)

**When NOT to use for loops in R:**
- Applying the same function to each element (use `map()` instead)
- Simple transformations of vectors (use vectorized operations)
- Filtering or summarizing data (use `dplyr` verbs)

### Example: Analyzing Multiple Counties

```{r county-analysis-setup}
# Create sample county data
county_data <- tibble(
  county = c("Alameda", "Orange", "San Diego", "Los Angeles", "Riverside"),
  population = c(1671329, 3186989, 3298634, 10014009, 2418185),
  median_income = c(94588, 84223, 70824, 68044, 63948),
  unemployment_rate = c(0.034, 0.029, 0.031, 0.041, 0.038)
)

print(county_data)
```

```{r for-loop-analysis}
# Initialize storage for results
county_analysis <- tibble(
  county = character(),
  economic_score = numeric(),
  risk_level = character()
)

# Loop through each county
for (i in 1:nrow(county_data)) {
  # Get current county data
  current_county <- county_data[i, ]
  
  # Calculate economic health score (higher income = better, lower unemployment = better)
  # Normalize income (divide by 100,000) and subtract unemployment rate
  economic_score <- (current_county$median_income / 100000) - (current_county$unemployment_rate * 10)
  
  # Determine risk level
  if (economic_score > 0.5) {
    risk_level <- "Low Risk"
  } else if (economic_score > 0) {
    risk_level <- "Moderate Risk"
  } else {
    risk_level <- "High Risk"
  }
  
  # Add to results
  county_analysis <- county_analysis %>%
    add_row(
      county = current_county$county,
      economic_score = economic_score,
      risk_level = risk_level
    )
}

# Display results
county_analysis %>%
  arrange(desc(economic_score)) %>%
  mutate(economic_score = round(economic_score, 2))
```

## The map() Family: Functional Programming

The `map()` functions from `purrr` are perfect when you want to:

- Apply the same function to multiple inputs
- Work with lists or vectors efficiently
- Write more concise, readable code

### Basic map() Usage

```{r map-basics}
# Apply our housing affordability function to multiple scenarios
income_scenarios <- c(50000, 75000, 100000, 125000)
rent_scenarios <- c(1500, 2000, 2500, 3000)

# Using map to apply function to multiple incomes (with fixed rent)
affordability_by_income <- map(
  .x = income_scenarios,  # .x = the vector we want to iterate over
  .f = ~ assess_housing_affordability(.x, median_rent = 2000)  # .f = the function to apply to each element
  # The ~ creates a formula/anonymous function
  # Inside the formula, .x refers to each individual element from the input vector
  # So this calls assess_housing_affordability(50000, 2000), then (75000, 2000), etc.
)

# Extract just the categories for easy viewing
# map_chr() is like map() but returns a character vector instead of a list
# It takes each result from affordability_by_income (which are lists) and extracts the $category element
income_categories <- map_chr(affordability_by_income, ~ .x$category)
names(income_categories) <- paste0("$", scales::comma(income_scenarios))
print(income_categories)
```


# Part IV: Advanced Techniques

## Error Handling in Functions

When working with real data, things can go wrong. Here's how to make your functions robust:

```{r error-handling}
safe_calculate_rate <- function(numerator, denominator) {
  # Check for invalid inputs
  if (!is.numeric(numerator) || !is.numeric(denominator)) {
    stop("Both inputs must be numeric")
  }
  
  if (denominator == 0) {
    warning("Denominator is zero, returning NA")
    return(NA)
  }
  
  if (denominator < 0) {
    warning("Negative denominator detected")
  }
  
  rate <- numerator / denominator
  return(rate)
}

# Test error handling
safe_calculate_rate(100, 200)  # Normal case
safe_calculate_rate(100, 0)    # Division by zero

# Demonstrate error handling (wrapped in try() so it doesn't break the document)
cat("Testing with invalid input:\n")
result <- try(safe_calculate_rate("100", 200), silent = TRUE)
if (inherits(result, "try-error")) {
  cat("Error caught:", attr(result, "condition")$message, "\n")
}
```

# Part V: Functions and Iteration in Machine Learning

The skills you've learned in this lab are the foundation for machine learning workflows. Here's a preview of how these concepts will be essential in upcoming weeks:

### Cross-Validation and Model Selection
```{r preview-cv, eval=FALSE}
# You'll use map() for k-fold cross-validation
cv_folds <- 1:5
cv_results <- map(cv_folds, function(fold) {
  # Split data, train model, evaluate
  train_elastic_net(train_data, test_data)
})

# Combine results to select best model
best_model <- cv_results %>% 
  map_dbl(~ .x$rmse) %>% 
  which.min()
```

### Hyperparameter Tuning
```{r preview-tuning, eval=FALSE}
# Grid search using the iteration skills from today
alpha_values <- seq(0, 1, by = 0.1)
tuning_results <- map(alpha_values, function(alpha) {
  evaluate_elastic_net(alpha = alpha, data = train_data)
})
```

### Robust ML Pipelines
```{r preview-robust, eval=FALSE}
# Error handling prevents crashes during model training
safe_xgboost <- possibly(train_xgboost, otherwise = NULL)
model_results <- map(param_combinations, safe_xgboost)
```

**Key insight**: Machine learning is fundamentally about applying the same analysis (data pipeline, model training and evaluation) across many different scenarios (hyperparameters, cross-validation folds, feature sets). The iteration skills you've learned today make this efficient and reproducible.



# Part V: Lab Assignment

## Assignment: Labor Action Tracker Code Review and Optimization

**Context**: In your LAT pipeline, you're likely encountering repetitive code that could benefit from functions and iteration.

### Part A: Team Collaboration (Work with your teammates)

**Step 1: Code Review**
Consult with your team and review your current Labor Action Tracker code together. Look for:

- **Repetitive analysis patterns**: Are you applying the same analysis to different states, time periods, or action types?
- **Copy-paste code**: Do you have similar code blocks with only small variations?
- **Manual processes**: Are you manually filtering, summarizing, or visualizing data in repetitive ways?
- **Hard-coded values**: Are there magic numbers or repeated calculations that could be parameterized?

**Step 2: Identify Opportunities**
As a team, identify at least 3-5 instances where functions or iteration could improve your workflow. Examples might include:

- Analyzing labor actions by state (same analysis, different state filter)
- Creating summary statistics for different time periods
- Generating similar plots for different variables
- Cleaning or processing data in repetitive ways
- Calculating metrics across different subgroups

**Step 3: Divide and Conquer**
Assign each team member 1-2 different optimization tasks. **Each person should work on different functions/iterations** - no duplicate work!

### Part B: Individual Implementation (Work independently)

**Your Task**: Implement your assigned function(s) or iteration pattern(s) to solve one of the repetitive code problems your team identified.

**Requirements**:

1. **Create at least one function** that eliminates repetitive code from your team's LAT analysis
2. **Use iteration if applicable** (and it very often will be) to apply your function across multiple scenarios
3. **Include error handling** where appropriate (consider using `possibly()` if some iterations might fail)
4. **Document your function** with clear comments explaining parameters and return values
5. **Test your function** with simulated data or real LAT data (ideally both) to show it works

**Example Structure**:
```{r your-function, eval=FALSE}
# Example: Function to count actions by state
count_state_actions <- function(lat_data, state_name) {
  state_data <- lat_data %>%
    filter(state == state_name)
  
  total_actions <- nrow(state_data)
  
  return(total_actions)
}

# Use a for loop to apply across multiple states
target_states <- c("California", "Texas", "New York", "Florida")
state_counts <- c()  # Initialize empty vector

for(i in 1:length(target_states)) {
  state_name <- target_states[i]
  count <- count_state_actions(lat_data, state_name)
  state_counts[i] <- count
  
  cat("State:", state_name, "- Actions:", count, "\n")
}

# Name the results
names(state_counts) <- target_states
```

### Part C: Individual Submission

**What to submit**:

1. **Your optimized code** with the function(s) and iteration you implemented (10 points)
2. **Before/after comparison** showing the repetitive code you replaced (5 points)
3. **Brief explanation** (2-3 sentences) of how your function improves your team's workflow (5 points)

**Submission format**: Submit your individual `.qmd` file with your specific optimization work.


**Due**: November 10, 2025


