---
title: "Machine Learning Workflows"
subtitle: "From Features to Predictions with caret"
author: "Rod Albuyeh"
date: today
format: pdf
execute:
  echo: true
  eval: true
  warning: false
  error: false
  cache: true
  tidy: true
editor: 
  markdown: 
    wrap: 72
---

# Introduction

This lab bridges the gap between machine learning theory and practice. In Lecture 7, we learned about supervised learning, the bias-variance tradeoff, and evaluation metrics. Now we'll apply these concepts to real data using the `caret` package—one of the most popular frameworks for machine learning in R.

## Learning Objectives

By the end of this lab, you will be able to:

1. **Set up ML workflows** using the caret package
2. **Engineer features** from raw data to improve model performance
3. **Implement proper data preprocessing** within a validation framework
4. **Train and evaluate models** using cross-validation
5. **Compare multiple models** systematically
6. **Interpret classification metrics** in policy contexts

## Why caret?

The `caret` package (Classification And REgression Training) provides:

- **Unified interface**: Train 200+ different models with the same syntax
- **Built-in preprocessing**: Scaling, centering, imputation, etc.
- **Cross-validation**: Automated k-fold CV and other resampling methods
- **Parallel processing**: Speed up training on multiple cores
- **Feature selection**: Built-in methods for reducing dimensionality
- **Consistent evaluation**: Same metrics across all models

Think of caret as the "tidyverse for machine learning"—it standardizes what would otherwise be a mess of different packages and syntaxes.

# Setup

```{r}
# Essential libraries
library(tidyverse)      # Data manipulation
library(caret)          # ML workflows
library(readxl)         # Reading Excel files
library(here)           # File paths
library(janitor)        # Cleaning names
library(scales)         # Pretty formatting

# For reproducibility
set.seed(452)

# Configure caret to use multiple cores (speeds up cross-validation)
library(doParallel)
cl <- makePSOCKcluster(2)  # Use 2 cores (adjust based on your computer)
registerDoParallel(cl)
```

# Part 1: Understanding the caret Workflow

## The Big Picture

Every machine learning workflow follows these steps:

1. **Load and clean data**
2. **Split into train/val/test sets**
3. **Define preprocessing steps**
4. **Train model(s)**
5. **Evaluate and compare**
6. **Make final predictions**

caret helps us do all of this in a standardized, reproducible way. Let's see it in action with a simple example first.

## A Simple Example: Policy Implementation Success

Let's walk through a complete workflow with simpler data so you can see how all the pieces fit together.

```{r}
# Create sample policy data
policy_data <- tibble(
  city_id = 1:500,
  population = rlnorm(500, log(100000), 1),
  budget_per_capita = rlnorm(500, log(1000), 0.5),
  staff_experience = runif(500, 1, 20),
  community_support = runif(500, 0, 100),
  prior_success = sample(c("Yes", "No"), 500, replace = TRUE, prob = c(0.3, 0.7))
) %>%
  mutate(
    # Create outcome based on predictors
    success_prob = plogis(-2 + 
                         0.0001 * budget_per_capita + 
                         0.1 * staff_experience + 
                         0.02 * community_support +
                         ifelse(prior_success == "Yes", 0.8, 0)),
    success = factor(
      ifelse(runif(500) < success_prob, "Success", "Failure"),
      levels = c("Failure", "Success")  # "Failure" is reference level
    )
  ) %>%
  select(-success_prob)  # Remove true probability

# Look at our data
glimpse(policy_data)
```

### Step 1: Train-Test Split

```{r}
# Create stratified split (maintains class proportions)
trainIndex <- createDataPartition(
  y = policy_data$success,    # Outcome variable
  p = 0.75,                    # 75% for training
  list = FALSE                 # Return indices, not list
)

# Split the data
train_data <- policy_data[trainIndex, ]
test_data <- policy_data[-trainIndex, ]

# Check class balance
cat("Training set class distribution:\n")
prop.table(table(train_data$success))

cat("\nTest set class distribution:\n")
prop.table(table(test_data$success))
```

::: callout-note
## Why Stratified Splitting?

`createDataPartition()` ensures that class proportions are maintained in both train and test sets. This is crucial for imbalanced data—if we have 70% Failures and 30% Successes overall, we want roughly the same split in both sets.
:::

### Step 2: Define Preprocessing

```{r}
# Set up preprocessing steps
preProc_values <- preProcess(
  train_data %>% select(-city_id, -success),  # Only use training data!
  method = c("center",      # Subtract mean
             "scale",       # Divide by SD
             "nzv")         # Remove near-zero variance predictors
)

# Apply to training data
train_transformed <- predict(preProc_values, train_data)

# Apply SAME transformation to test data (no data leakage!)
test_transformed <- predict(preProc_values, test_data)

# Show what happened
cat("Original training data (first 3 rows):\n")
print(head(train_data %>% select(budget_per_capita, staff_experience), 3))

cat("\nTransformed training data (first 3 rows):\n")
print(head(train_transformed %>% select(budget_per_capita, staff_experience), 3))

cat("\nNote: Values are now centered (mean ≈ 0) and scaled (SD ≈ 1)")
```

::: callout-important
## Preventing Data Leakage

Notice that we:

1. Fit `preProcess()` using **only training data**
2. Apply the same transformation to test data using `predict()`

This prevents test data from influencing our preprocessing!
:::

### Step 3: Set Up Cross-Validation

```{r}
# Define cross-validation strategy
ctrl <- trainControl(
  method = "cv",              # Use k-fold cross-validation
  number = 5,                 # 5 folds
  summaryFunction = twoClassSummary,  # Classification metrics
  classProbs = TRUE,          # Need probabilities for ROC
  savePredictions = "final"   # Save predictions for analysis
)
```

**What does this control object do?**

- **Method**: k-fold cross-validation (cv)
- **Number of folds**: 5
- **Metrics computed**: ROC-AUC, Sensitivity, Specificity
- **Class probabilities**: Enabled for ROC curve calculation
- **Predictions saved**: Final model predictions stored for later analysis

### Step 4: Train a Model

```{r}
# Train a logistic regression model
model_fit <- train(
  success ~ budget_per_capita + staff_experience + 
            community_support + prior_success,
  data = train_transformed,
  method = "glm",                    # Generalized linear model (logistic regression)
  family = "binomial",               # For binary classification
  trControl = ctrl,                  # Use our CV setup
  metric = "ROC"                     # Optimize for ROC-AUC
)

# Look at results
print(model_fit)
```

### Step 5: Evaluate Performance

```{r}
# Get predictions on test set
test_preds <- predict(model_fit, newdata = test_transformed, type = "prob")
test_class <- predict(model_fit, newdata = test_transformed)

# Create confusion matrix
conf_mat <- confusionMatrix(
  data = test_class,
  reference = test_transformed$success,
  positive = "Success"  # Define which class is "positive"
)

print(conf_mat)
```

Let's break down what these metrics mean:

```{r}
# Extract key metrics
accuracy <- conf_mat$overall['Accuracy']
sensitivity <- conf_mat$byClass['Sensitivity']  # Recall / True Positive Rate
specificity <- conf_mat$byClass['Specificity']  # True Negative Rate
precision <- conf_mat$byClass['Precision']      # Positive Predictive Value

cat("Model Performance on Test Set:\n")
cat("================================\n")
cat("Accuracy:    ", round(accuracy, 3), "- Overall % correct\n")
cat("Sensitivity: ", round(sensitivity, 3), "- Of actual Successes, % we caught\n")
cat("Specificity: ", round(specificity, 3), "- Of actual Failures, % we caught\n")
cat("Precision:   ", round(precision, 3), "- Of predicted Successes, % were correct\n")
```

### Visualizing Performance

```{r}
# ROC curve
library(pROC)

roc_obj <- roc(
  response = test_transformed$success,
  predictor = test_preds$Success
)

# Plot ROC curve
plot(roc_obj, 
     main = "ROC Curve for Policy Success Prediction",
     col = "steelblue",
     lwd = 2)
abline(a = 0, b = 1, lty = 2, col = "red")  # Random classifier line

# Add AUC to plot
text(0.5, 0.3, 
     paste("AUC =", round(auc(roc_obj), 3)),
     cex = 1.2)
```

::: callout-note
## Interpreting ROC-AUC

- **AUC = 0.5**: Model is no better than random guessing
- **AUC = 0.6-0.7**: Acceptable for very hard problems or poor quality data—still better than nothing!
- **AUC = 0.7-0.8**: Good discrimination for most policy applications
- **AUC = 0.8-0.9**: Very good discrimination
- **AUC > 0.9**: Excellent discrimination (be suspicious—might be data leakage!)

**Policy reality**: Human behavior is inherently noisy. An AUC of 0.65 predicting labor strikes or program enrollment might be genuinely useful even though it wouldn't impress in a Kaggle competition. Context matters—if your baseline is random guessing (0.5), even 0.6 represents real predictive signal that can inform better decisions.
:::


# Part 2: Feature Engineering

Feature engineering is the process of creating new variables from existing ones to improve model performance. Good features can make the difference between a mediocre model and a great one.

## Types of Feature Engineering

### 1. Transformations

```{r transformations-plot}
#| fig.height: 6
#| fig.width: 8

# Create demonstration data with skewed variables
demo_data <- tibble(
  income = rlnorm(1000, log(50000), 0.8),
  population = rlnorm(1000, log(100000), 1)
)

# Compare distributions before and after log transformation
par(mfrow = c(2, 2))

hist(demo_data$income, main = "Income (Original)", 
     xlab = "Income", col = "lightblue")
hist(log(demo_data$income), main = "Income (Log Transformed)",
     xlab = "log(Income)", col = "lightgreen")

hist(demo_data$population, main = "Population (Original)",
     xlab = "Population", col = "lightblue")
hist(log(demo_data$population), main = "Population (Log Transformed)",
     xlab = "log(Population)", col = "lightgreen")

# Reset graphics parameters
par(mfrow = c(1, 1))
```

**When to use log transformations:**
- Variables that span multiple orders of magnitude (e.g., income, population)
- Variables where proportional changes matter more than absolute changes
- When you see severe right skew in histograms

### 2. Interaction Terms

```{r}
# Interaction: effect of one variable depends on another
policy_data_interact <- train_data %>%
  mutate(
    # Maybe budget per capita matters more in larger cities?
    budget_pop_interact = budget_per_capita * log(population),
    
    # Maybe staff experience matters more with community support?
    staff_support_interact = staff_experience * community_support
  )

# Train model with interactions
model_interact <- train(
  success ~ budget_per_capita + staff_experience + 
            community_support + prior_success +
            budget_pop_interact + staff_support_interact,
  data = policy_data_interact,
  method = "glm",
  family = "binomial",
  trControl = ctrl,
  metric = "ROC"
)

cat("Model without interactions - CV ROC:", 
    round(max(model_fit$results$ROC), 3), "\n")
cat("Model with interactions - CV ROC:", 
    round(max(model_interact$results$ROC), 3), "\n")
```

### 3. Binning Continuous Variables

Sometimes it helps to group continuous variables into categories:

```{r}
# Create bins for staff experience
policy_data_binned <- train_data %>%
  mutate(
    experience_level = cut(
      staff_experience,
      breaks = c(0, 5, 10, 20),
      labels = c("Junior", "Mid-Level", "Senior"),
      include.lowest = TRUE
    )
  )

# Compare with continuous version
model_binned <- train(
  success ~ budget_per_capita + experience_level + 
            community_support + prior_success,
  data = policy_data_binned,
  method = "glm",
  family = "binomial",
  trControl = ctrl,
  metric = "ROC"
)

cat("Continuous experience - CV ROC:", 
    round(max(model_fit$results$ROC), 3), "\n")
cat("Binned experience - CV ROC:", 
    round(max(model_binned$results$ROC), 3), "\n")
```

::: callout-note
## Binning: When It Helps and Hurts

**Helps when:**

- Non-linear relationships that are approximately step-wise
- Interpretability is crucial (easier to explain "senior staff" than "staff with 15.3 years experience")
- You have domain knowledge about meaningful thresholds

**Hurts when:**

- Relationships are actually smooth and continuous
- You lose information by grouping
- Bins are arbitrary and not grounded in domain knowledge
:::

### 4. Domain-Specific Features

The best features often come from domain knowledge. For policy data, consider:

```{r}
# Create policy-relevant features
policy_data_domain <- train_data %>%
  mutate(
    # Resources per capita (maybe more meaningful than absolute budget)
    resources_per_capita = budget_per_capita / population * 1000000,
    
    # Experience-to-complexity ratio
    staff_readiness = staff_experience / log(population),
    
    # High-risk category: low support AND low experience
    high_risk = factor(
      ifelse(community_support < 50 & staff_experience < 10, "High Risk", "Lower Risk")
    )
  )

glimpse(policy_data_domain)
```


# Part 3: Working with Real Data - Credit Default Prediction

Now let's apply everything we've learned to real data: predicting credit card default using the `Default` dataset from the `ISLR` package. This is a classic dataset in statistical learning.

## Loading and Exploring the Data

```{r}
# Load the ISLR package (install if needed)
# install.packages("ISLR")
library(ISLR)

# Load the Default dataset
data(Default)

# Look at the data
glimpse(Default)

# Check the dataset
cat("Dataset dimensions:", nrow(Default), "observations,", ncol(Default), "variables\n\n")

# View first few rows
head(Default)

# Check outcome distribution
cat("\nOutcome distribution:\n")
table(Default$default)
cat("\nProportions:\n")
prop.table(table(Default$default))
```

::: callout-important
## Class Imbalance

Notice that our outcome is imbalanced (not 50-50). This is common in real-world data. We'll need to keep this in mind when:

- Interpreting accuracy (might be misleading)
- Choosing evaluation metrics (precision-recall curves often better than ROC)
- Deciding on interventions (false positives vs false negatives have different costs)
:::

## Feature Engineering for Credit Default

```{r}
# Create features specific to credit risk
credit_features <- Default %>%
  mutate(
    # Financial ratio features
    balance_income_ratio = balance / income,
    
    # Log transformations for skewed variables
    log_balance = log1p(balance),  # log1p(x) = log(x + 1), handles zeros
    log_income = log1p(income),
    
    # Categorical indicators
    is_student = factor(student),
    
    # Risk categories based on balance
    balance_category = cut(
      balance,
      breaks = c(0, 500, 1000, 2000, Inf),
      labels = c("Low", "Medium", "High", "Very High"),
      include.lowest = TRUE
    ),
    
    # Income categories
    income_category = cut(
      income,
      breaks = quantile(income, probs = c(0, 0.25, 0.5, 0.75, 1)),
      labels = c("Q1", "Q2", "Q3", "Q4"),
      include.lowest = TRUE
    ),
    
    # High-risk indicator: high balance AND low income
    high_risk_flag = factor(
      ifelse(balance > median(balance) & income < median(income), 
             "High Risk", "Lower Risk")
    )
  )

# Look at our engineered features
glimpse(credit_features %>% select(default, balance_income_ratio, log_balance, 
                                  balance_category, high_risk_flag))
```

## Model Training with Cross-Validation

```{r}
# Select features for modeling
modeling_data <- credit_features %>%
  select(
    default,
    balance,
    income,
    student,
    log_balance,
    log_income,
    balance_income_ratio,
    balance_category,
    high_risk_flag
  ) %>%
  drop_na()  # Remove any remaining missing values

# Create train-test split (stratified by outcome)
set.seed(452)
trainIndex <- createDataPartition(
  modeling_data$default,
  p = 0.75,
  list = FALSE
)

train_credit <- modeling_data[trainIndex, ]
test_credit <- modeling_data[-trainIndex, ]

# Set up cross-validation with appropriate metrics for imbalanced data
ctrl_credit <- trainControl(
  method = "cv",
  number = 5,
  summaryFunction = twoClassSummary,
  classProbs = TRUE,
  savePredictions = "final"
)

# Train logistic regression
model_credit <- train(
  default ~ .,
  data = train_credit,
  method = "glm",
  family = "binomial",
  trControl = ctrl_credit,
  metric = "ROC",
  preProcess = c("center", "scale", "nzv")  # Preprocessing within caret
)

# View results
print(model_credit)
```

## Evaluating Model Performance

```{r}
# Predictions on test set
test_preds_credit <- predict(model_credit, newdata = test_credit, type = "prob")
test_class_credit <- predict(model_credit, newdata = test_credit)

# Confusion matrix
conf_mat_credit <- confusionMatrix(
  data = test_class_credit,
  reference = test_credit$default,
  positive = "Yes"  # "Yes" means defaulted
)

print(conf_mat_credit)
```

### Visualizing Results

```{r}
# ROC Curve
roc_credit <- roc(
  response = test_credit$default,
  predictor = test_preds_credit$Yes
)

plot(roc_credit,
     main = "ROC Curve: Predicting Credit Default",
     col = "darkblue",
     lwd = 2)
abline(a = 0, b = 1, lty = 2, col = "red")
text(0.5, 0.3,
     paste("AUC =", round(auc(roc_credit), 3)),
     cex = 1.2)
```

```{r}
# Variable importance
var_imp <- varImp(model_credit)
plot(var_imp, main = "Feature Importance for Default Prediction")
```

::: callout-note
## Interpreting Variable Importance

In logistic regression, variable importance is based on the absolute value of the coefficients. Higher values mean the feature has more influence on predictions.

**Financial policy insight**: Understanding which features matter most helps:

- Guide data collection efforts (which customer info is most predictive?)
- Identify key risk factors for intervention (target outreach to high-risk customers)
- Explain model decisions to stakeholders (regulators, customers, executives)
:::

## Model Comparison

Let's compare different modeling approaches:

```{r}
# Model 1: Simple logistic regression (already trained above)
# Model 2: Regularized logistic regression (Ridge)
model_ridge <- train(
  default ~ .,
  data = train_credit,
  method = "glmnet",
  family = "binomial",
  trControl = ctrl_credit,
  metric = "ROC",
  preProcess = c("center", "scale", "nzv"),
  tuneGrid = expand.grid(
    alpha = 0,  # Ridge regression (alpha = 0)
    lambda = seq(0.001, 0.1, length = 20)
  )
)

# Model 3: Elastic Net (combination of Ridge and LASSO)
model_enet <- train(
  default ~ .,
  data = train_credit,
  method = "glmnet",
  family = "binomial",
  trControl = ctrl_credit,
  metric = "ROC",
  preProcess = c("center", "scale", "nzv"),
  tuneGrid = expand.grid(
    alpha = seq(0, 1, length = 5),  # Mix of L1 and L2
    lambda = seq(0.001, 0.1, length = 20)
  )
)

# Compare models using resamples
results <- resamples(list(
  Logistic = model_credit,
  Ridge = model_ridge,
  ElasticNet = model_enet
))

# Summary statistics
summary(results)
```

```{r}
# Visualize comparison
bwplot(results, metric = "ROC")
```

```{r}
# Test set performance for each model
test_results <- tibble(
  Model = c("Logistic", "Ridge", "Elastic Net"),
  AUC = c(
    auc(roc(test_credit$default, predict(model_credit, test_credit, type = "prob")$Yes)),
    auc(roc(test_credit$default, predict(model_ridge, test_credit, type = "prob")$Yes)),
    auc(roc(test_credit$default, predict(model_enet, test_credit, type = "prob")$Yes))
  )
) %>%
  arrange(desc(AUC))

print(test_results)
```


# Part 4: Production Considerations

## Saving and Loading Models

```{r}
# Save your trained model
saveRDS(model_credit, here("data", "processed", "credit_default_model.rds"))

# Load it back
# loaded_model <- readRDS(here("data", "processed", "credit_default_model.rds"))

# Make predictions with loaded model
# new_predictions <- predict(loaded_model, newdata = new_data, type = "prob")
```

## Model Documentation

Every production model needs documentation:

```{r}
# Create a model card
model_card <- list(
  model_name = "Credit Default Risk Classifier",
  version = "1.0",
  date_trained = Sys.Date(),
  training_data = "ISLR Default dataset (n = 10,000 customers)",
  algorithm = "Logistic Regression",
  features = names(train_credit)[-1],  # All except outcome
  performance = list(
    cv_auc = max(model_credit$results$ROC),
    test_auc = auc(roc_credit),
    accuracy = conf_mat_credit$overall['Accuracy'],
    sensitivity = conf_mat_credit$byClass['Sensitivity'],
    specificity = conf_mat_credit$byClass['Specificity']
  ),
  preprocessing = "Centering, Scaling, NZV removal",
  notes = "Model trained for educational purposes. Not for production use without validation."
)

# Save model card
saveRDS(model_card, here("data", "processed", "model_card.rds"))

# Print it nicely
cat("=== MODEL CARD ===\n")
cat("Model:", model_card$model_name, "\n")
cat("Version:", model_card$version, "\n")
cat("Trained:", as.character(model_card$date_trained), "\n")
cat("Algorithm:", model_card$algorithm, "\n")
cat("\nPerformance:\n")
cat("  CV AUC:", round(model_card$performance$cv_auc, 3), "\n")
cat("  Test AUC:", round(model_card$performance$test_auc, 3), "\n")
cat("  Accuracy:", round(model_card$performance$accuracy, 3), "\n")
```

## Monitoring for Drift

In production, you need to monitor:

1. **Data drift**: Are input distributions changing?
2. **Concept drift**: Is the relationship between features and outcome changing?
3. **Performance drift**: Is model accuracy degrading?

```{r drift-check-plot}
#| fig.height: 4
#| fig.width: 8

# Example: Check if distributions are similar between train and test
# Compare balance distributions
par(mfrow = c(1, 2))
hist(train_credit$balance, main = "Training Set",
     xlab = "Credit Card Balance", col = "lightblue", breaks = 30)
hist(test_credit$balance, main = "Test Set",
     xlab = "Credit Card Balance", col = "lightgreen", breaks = 30)

# Reset graphics parameters
par(mfrow = c(1, 1))
```

```{r}
# Statistical test for distribution difference
ks_test <- ks.test(train_credit$balance, test_credit$balance)
cat("Kolmogorov-Smirnov test p-value:", round(ks_test$p.value, 4), "\n")
cat("Interpretation:", 
    ifelse(ks_test$p.value > 0.05, 
           "Distributions are similar (good!)",
           "Distributions differ (potential drift!)"))
```

# Lab Assignment: Build Your Complete ML Workflow

## Overview

Your task is to build a complete machine learning workflow for a classification problem of your choice, incorporating everything you've learned in this lab. You may use:

- The `Default` dataset (credit default prediction) - what we used in class
- Your **Labor Action Tracker data** (strike vs protest classification) - recommended!
- Another classification dataset that interests you (get approval first)

## Requirements

### Part 1: Data Preparation and Feature Engineering (8 points)

1. **Load and clean your chosen data** (2 points)
   - Handle missing values appropriately
   - Create proper train/test splits (consider stratification!)
   - Document your decisions

2. **Engineer at least 5 meaningful features** (6 points)
   - At least one transformation (e.g., log, polynomial)
   - At least one interaction term
   - At least one domain-specific feature
   - At least one feature inspired by your skim of the Kuhn book.
   - Explain the reasoning behind each feature

### Part 2: Model Development (10 points)

1. **Set up proper validation** (2 points)
   - Choose appropriate validation strategy
   - Justify your choice

2. **Train at least 2 different models** (4 points)
   - Use logistic regression as baseline
   - Try at least one other approach (Ridge, Elastic Net, or something else)
   - Use caret's unified interface

3. **Implement proper preprocessing** (2 points)
   - Scale/center where appropriate
   - Handle categorical variables correctly
   - Avoid data leakage

4. **Compare models systematically** (2 points)
   - Use cross-validation results
   - Test on held-out data
   - Visualize comparisons

### Part 3: Evaluation and Interpretation (7 points)

1. **Compute comprehensive metrics** (3 points)
   - ROC-AUC
   - Precision, Recall, F1
   - Confusion matrix with interpretation

2. **Create visualizations** (2 points)
   - ROC curve
   - Variable importance plot
   - At least one custom visualization

3. **Policy/Business interpretation** (2 points)
   - Which factors most strongly predict your outcome?
   - What are the policy or business implications?
   - Discuss costs of false positives vs false negatives in your context
   - How would you explain this model to a relevant stakeholder?

### Part 4: Documentation (5 points)

1. **Write a model card** (2 points)
   - Document your model's key characteristics
   - Include performance metrics
   - Note limitations and assumptions

2. **Code quality** (3 points)
   - Clear, readable code with comments
   - Reproducible (use set.seed())
   - Follows best practices from the course

## Submission Guidelines

- Submit both `.qmd` and `.pdf` files
- Include all code (but hide code with `echo: false` in final PDF where it improves readability)
- Keep the report focused: aim for 8-12 pages including visualizations
- Due: **November 17, 2025 at 11:59 PM**

## Tips for Success

1. **Start simple**: Get a baseline working first, then improve
2. **Feature engineering is key**: This is where you'll see the biggest gains
3. **Cross-validation is crucial**: Don't trust single train-test split results
4. **Think like a policy maker**: How would you actually use this model?
5. **Document as you go**: Don't wait until the end to write explanations

## Academic Integrity

If you use AI tools (ChatGPT, Claude, etc.):

- Include an appendix stating which tools you used
- Explain what help you requested
- Describe how you verified and modified AI-generated content
- Remember: You're responsible for understanding and explaining all code

## Grading Rubric

- **Feature Engineering (8 pts)**: Quality, creativity, and justification of features
- **Model Development (10 pts)**: Technical correctness, proper validation, comparison
- **Evaluation (7 pts)**: Comprehensive metrics, clear visualizations, thoughtful interpretation
- **Documentation (5 pts)**: Code quality, reproducibility, clear explanations
- **Total: 30 points**

---

**Good luck! Remember: The goal isn't to build a perfect model, but to demonstrate that you understand the ML workflow and can apply it thoughtfully to policy problems.**

```{r}
#| include: false

# Clean up parallel processing (if still active)
if (exists("cl") && inherits(cl, "cluster")) {
  tryCatch(
    stopCluster(cl),
    error = function(e) message("Cluster already stopped or invalid")
  )
}
```

