---
title: "Ethics in Data Science and Policy"
subtitle: "Principles, Practice, and Public Trust"
author: "Rod Albuyeh"
date: "10/20/2025"
format:
  revealjs:
    theme: [default, simple]
    css: styles.css
    slide-number: true
    progress: true
    overview: true
    controls: true
    transition: fade
    auto-animate: true
    incremental: true
    code-overflow: wrap
    code-line-numbers: false
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r setup, include=TRUE, echo=FALSE, results=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 8, fig.height = 6)
library(tidyverse)
library(scales)
library(patchwork)
```

## Today's Learning Goals {.smaller}

::: {.fragment}
By the end of this session, you'll be able to:
:::

::: {.fragment}
- **Apply** ethical frameworks to data science decisions
- **Evaluate** trade-offs between competing principles
- **Design** ethical safeguards in data projects
- **Communicate** ethical considerations to stakeholders
:::

::: {.fragment}
::: callout-note
**Key insight**: Ethics in data science isn't about following rules—it's about making principled decisions that respect human dignity and promote social good.
:::
:::

# Part I: Why Ethics Matter in Data Science

## The Stakes Have Never Been Higher {.smaller}

::: {.fragment}
**Data science decisions affect lives:**

- Credit scoring → Access to housing
- Risk assessment → Criminal justice outcomes
- Healthcare algorithms → Treatment access
- Facial recognition → Civil liberties
:::

::: {.fragment}
**Recent controversies:**

- Biased hiring algorithms
- Discriminatory lending practices
- Privacy breaches in health apps
- Surveillance overreach
:::

::: {.fragment}
::: callout-important
**Reality check**: Technical excellence without ethical consideration can amplify societal harms.
:::
:::

## Case Study: COMPAS Recidivism Algorithm {.smaller}

::: {.fragment}
**The setup:**

- Algorithm predicts likelihood of reoffending
- Used in sentencing and parole decisions
- Affects thousands of lives daily
:::

::: {.fragment}
**ProPublica's findings (2016):**

- Racially biased predictions
- False positive rate for Black defendants: 45%
- False positive rate for White defendants: 24%
:::

::: {.fragment}
**Key ethical questions:**
- What does "fairness" mean in this context?
- How do we balance accuracy vs. equity?
- Should algorithms make high-stakes decisions?
:::

# Part II: Ethical Frameworks in Practice

## Beyond Rules: A Principled Approach {.smaller}

::: {layout-ncol=2}
### Traditional Approaches
- **Rules-based (IRB)**
  + Slow to adapt
  + Binary decisions
  + Minimum standards
- **Ad hoc**
  + Inconsistent
  + Lacks accountability
  + Personal bias risk

### Modern Framework
- **Principle-based**
  + Flexible
  + Context-aware
  + Team-oriented
- **Outcome-focused**
  + Impact assessment
  + Stakeholder input
  + Continuous evaluation
:::

::: {.fragment}
**Key insight**: Good ethical practice requires both strong principles AND practical frameworks for applying them.
:::

## Core Ethical Principles in Data Science {.smaller}

::: {.fragment}
**1. Respect for Persons**
- Individual autonomy
- Informed consent
- Right to opt out
:::

::: {.fragment}
**2. Beneficence**
- Maximize benefits
- Minimize harm
- Consider indirect effects
:::

::: {.fragment}
**3. Justice**
- Fair distribution of benefits
- Protection of vulnerable groups
- Equal access to recourse
:::

::: {.fragment}
**4. Transparency**
- Explainable decisions
- Accessible documentation
- Clear communication
:::

## Ethical Frameworks Across Cultures {.smaller}

```{r}
#| echo: false
# Create a comparison of ethical frameworks
frameworks <- tibble(
  Framework = c("Belmont Report (US)", "FATML", "UK Data Ethics", "China TC260"),
  Principles = c(
    "Respect, Beneficence, Justice",
    "Responsibility, Explainability, Accuracy, Auditability, Fairness",
    "Transparency, Accountability, Fairness",
    "Harmony, Justice, Privacy, Safety, Collaboration"
  ),
  Context = c(
    "Research Ethics",
    "Machine Learning",
    "Government Data",
    "Information Systems"
  )
)

knitr::kable(frameworks)
```

::: {.fragment}
**Common themes:**
- Fairness and justice
- Transparency
- Responsibility
- Protection of rights
:::

# Part III: Practical Ethics in Data Projects

## The Ethics Canvas: A Tool for Teams {.smaller}

::: {.fragment}
**Step 1: Stakeholder Analysis**
- Who is affected?
- What are their interests?
- Where are potential conflicts?
:::

::: {.fragment}
**Step 2: Risk Assessment**
- Privacy risks
- Bias potential
- Unintended consequences
:::

::: {.fragment}
**Step 3: Mitigation Planning**
- Technical safeguards
- Process controls
- Monitoring systems
:::

::: {.fragment}
**Step 4: Documentation**
- Decision rationale
- Trade-off analysis
- Impact assessment
:::

## Interactive Exercise: Ethical Decision Making {.smaller}

::: {.fragment}
**Scenario**: City wants to use utility data to identify households for targeted social services.

**Your task (5 minutes):**
1. List stakeholders
2. Identify ethical concerns
3. Propose safeguards
4. Consider trade-offs
:::

::: {.fragment}
**Discussion points:**
- Privacy vs. public good
- Consent requirements
- Data security
- Fairness in targeting
:::

## Privacy: Beyond PII {.smaller}

::: {.fragment}
**Traditional approach:**
- Remove names, SSNs, addresses
- Aggregate data
- Basic anonymization
:::

::: {.fragment}
**Modern challenges:**
- Re-identification through linkage
- Behavioral fingerprints
- Location data risks
- Temporal patterns
:::

```{r}
#| echo: false
# Create example of how seemingly anonymous data can be identifying
anonymous_data <- tibble(
  ZIP = c("92101", "92101", "92101", "92101"),
  Age = c(45, 45, 45, 45),
  Income = c("High", "High", "High", "High"),
  Visits = c(12, 15, 8, 10)
)

knitr::kable(anonymous_data, 
             caption = "Even 'anonymous' data can identify individuals when combined")
```

## Bias and Fairness: A Framework {.smaller}

::: {layout-ncol=2}
### Types of Bias
- **Selection bias**
  + Non-representative data
  + Sampling issues
- **Measurement bias**
  + Poor data quality
  + Inconsistent collection
- **Algorithm bias**
  + Model assumptions
  + Feature selection
- **Deployment bias**
  + Context mismatch
  + Feedback loops

### Fairness Metrics
- **Group fairness**
  + Equal error rates
  + Demographic parity
- **Individual fairness**
  + Similar cases
  + Similar outcomes
- **Process fairness**
  + Transparent
  + Contestable
- **Outcome fairness**
  + Impact analysis
  + Long-term effects
:::

## Hypothetical Case: Predictive Policing {.smaller}

::: {.fragment}
**The system:**
- Predicts crime hotspots
- Allocates police resources
- Uses historical crime data
:::

::: {.fragment}
**Ethical concerns:**
- Reinforcing historical biases
- Over-policing communities
- Self-fulfilling predictions
:::

::: {.fragment}
**Mitigation strategies:**
- Community input
- Multiple data sources
- Regular bias audits
- Clear oversight
:::

# Part IV: Building Ethical Data Science Teams

## Team Composition and Roles {.smaller}

::: {.fragment}
**Core competencies needed:**
- Data ethics expertise
- Legal knowledge
- Domain understanding
- Technical skills
- Communication ability
:::

::: {.fragment}
**Key roles:**
- Ethics lead
- Privacy officer
- Domain experts
- Technical team
- Stakeholder liaison
:::

::: {.fragment}
**Team effectiveness factors:**
- Complementary skill sets
- Cross-disciplinary expertise
- Domain knowledge depth
- Professional experience breadth
:::

## Ethics Review Process {.smaller}

::: {.fragment}
**Stage 1: Planning**
- Impact assessment
- Stakeholder consultation
- Risk identification
:::

::: {.fragment}
**Stage 2: Development**
- Regular ethics reviews
- Bias testing
- Privacy checks
:::

::: {.fragment}
**Stage 3: Deployment**
- Monitoring plan
- Feedback mechanisms
- Update procedures
:::

::: {.fragment}
**Stage 4: Evaluation**
- Impact analysis
- Stakeholder feedback
- Lessons learned
:::

## Practical Tools for Ethical Practice {.smaller}

::: {.fragment}
**Documentation templates:**
- Ethics canvas
- Risk register
- Decision log
:::

::: {.fragment}
**Technical tools:**
- Fairness metrics
- Privacy-preserving techniques
- Bias detection
:::

::: {.fragment}
**Process tools:**
- Ethics review checklist
- Stakeholder mapping
- Impact assessment
:::

::: {.fragment}
**Communication tools:**
- Plain language summaries
- Visual decision trees
- Stakeholder briefs
:::

# Part V: Future Challenges and Opportunities

## Emerging Ethical Challenges {.smaller}

::: {.fragment}
**AI and Automation**
- Algorithmic accountability
- Human oversight
- Automation bias
:::

::: {.fragment}
**Privacy Evolution**
- Data sovereignty
- Right to be forgotten
- Privacy-preserving ML
:::

::: {.fragment}
**Global Governance**
- Cross-border data flows
- Cultural differences
- Regulatory alignment
:::

::: {.fragment}
**Environmental Impact**
- Computing resources
- Energy consumption
- Sustainability
:::

## Building Public Trust {.smaller}

::: {.fragment}
**Key elements:**
- Transparent processes
- Clear communication
- Meaningful oversight
- Public engagement
:::

::: {.fragment}
**Best practices:**
- Regular audits
- Public reporting
- Stakeholder feedback
- Continuous improvement
:::

::: {.fragment}
::: callout-note
**Remember**: Trust is earned through consistent ethical practice, not just good intentions.
:::
:::

## Next Steps {.smaller}

::: {.fragment}
**Important Dates:**

- **Assignment 2 Due**: Friday, October 31st at 11:59 PM
- Next week we move on to algorithmic thinking.
- Following that, we advance to core machine learning algorithms.
:::


## Questions?

::: {.fragment}
**Remember**: Ethics isn't a constraint on data science—it's a foundation for building better, more trustworthy systems.
:::
