---
title: "Algorithmic Thinking for Policy Data Science"
subtitle: "Designing Efficient Approaches"
author: "Rod Albuyeh"
date: "11/03/2025"
format:
  revealjs:
    theme: [default, simple]
    css: styles.css
    slide-number: true
    progress: true
    overview: true
    controls: true
    transition: fade
    auto-animate: true
    incremental: true
    code-overflow: wrap
    code-line-numbers: false
    code-block-height: 300px
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r setup, include=TRUE, echo=FALSE, results=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 8, fig.height = 6)
library(tidyverse)
library(scales)
library(microbenchmark)
library(patchwork)
```

## Today's Learning Goals {.smaller}

::: {.fragment}
By the end of this session, you'll be able to:
:::

::: {.fragment}
- **Distinguish** between algorithms and their implementations
- **Analyze** computational complexity and its policy implications
- **Design** efficient approaches to common policy problems
- **Evaluate** trade-offs between different algorithmic strategies
:::

::: {.fragment}
::: callout-note
**Key insight**: In policy work, the choice of algorithm affects not just performance, but also fairness, transparency, and scalability of solutions.
:::
:::

# Part I: What Is Algorithmic Thinking?

## Algorithm vs. Implementation: A Policy Example {.smaller}

::: {.fragment}
**Scenario**: City needs to assign students to schools based on preferences and capacity
:::

::: columns
::: {.column width="50%"}
### The Algorithm (Recipe)
1. Collect student preferences
2. Check school capacities  
3. Apply matching rules
4. Handle conflicts fairly
5. Produce final assignments
:::

::: {.column width="50%"}
### The Implementation (Code)
::: {.fragment}
```r
assign_students <- function(student_prefs, school_caps) {
  assignments <- tibble()
  
  for(i in 1:nrow(student_prefs)) {
    # Find best available school
    # Check capacity constraints
    # Make assignment
  }
  
  return(assignments)
}
```
:::
:::
:::

::: {.fragment}
**Key distinction**: The algorithm is the logical approach; the implementation is how we code it in a specific language.
:::

## Why Algorithmic Thinking Matters in Policy {.smaller}

::: {.fragment}
**Real-world stakes:**
- **Efficiency**: Can we process 100,000 applications in reasonable time?
- **Fairness**: Does our approach treat all applicants equally?
- **Transparency**: Can stakeholders understand how decisions are made?
- **Scalability**: Will this work as the city grows?
:::

::: {.fragment}
**Policy examples where algorithms matter:**
- Vaccine distribution during emergencies
- Emergency response resource allocation
- Benefit eligibility determination
- Traffic light optimization
:::

::: {.fragment}
::: callout-important
**Remember**: In policy contexts, "good enough" algorithms that are transparent and fair often beat "optimal" algorithms that are opaque.
:::
:::

# Part II: Measuring Algorithm Quality

## The Three Dimensions of Algorithm Quality {.smaller}

::: {layout-ncol=3}
### Correctness
- Does it solve the problem?
- Are results accurate?
- Does it handle edge cases?

### Efficiency
- **Time**: How long does it take?
- **Space**: How much memory?
- **Resources**: What does it cost?

### Maintainability
- Can others understand it?
- Is it easy to modify?
- Can we audit decisions?
:::

::: {.fragment}
**Policy reality**: We often trade some efficiency for transparency and maintainability.
:::

## Case Study: Finding the Maximum Value {.smaller}

Let's compare different approaches to a simple problem: finding the largest value in a dataset.

::: {.fragment}
**Policy context**: Finding the county with the highest unemployment rate for targeted intervention.
:::

```{r}
# Create sample unemployment data
set.seed(42)
unemployment_data <- tibble(
  county = paste("County", 1:1000),
  unemployment_rate = runif(1000, 0.02, 0.15)  # 2% to 15%
)

# Show first few rows
head(unemployment_data, 3)
```

## Approach 1: The Straightforward Method {.smaller}

```{r}
# Method 1: Simple and clear
find_max_simple <- function(data) {
  max_rate <- data$unemployment_rate[1]  # Start with first value
  max_county <- data$county[1]
  
  for(i in 2:nrow(data)) {
    if(data$unemployment_rate[i] > max_rate) {
      max_rate <- data$unemployment_rate[i]
      max_county <- data$county[i]
    }
  }
  
  return(list(county = max_county, rate = max_rate))
}

# Test it
result1 <- find_max_simple(unemployment_data)
cat("Highest unemployment:", result1$county, "at", 
    scales::percent(result1$rate, accuracy = 0.1))
```

::: {.fragment}
**Pros**: Easy to understand, memory efficient, straightforward approach
**Cons**: Requires writing a loop, more verbose than built-in functions
:::

## Approach 2: Using Built-in Functions {.smaller}

```{r}
# Method 2: Leverage R's optimized functions
find_max_builtin <- function(data) {
  max_idx <- which.max(data$unemployment_rate)
  return(list(
    county = data$county[max_idx],
    rate = data$unemployment_rate[max_idx]
  ))
}

# Test it
result2 <- find_max_builtin(unemployment_data)
cat("Highest unemployment:", result2$county, "at", 
    scales::percent(result2$rate, accuracy = 0.1))
```

::: {.fragment}
**Pros**: Concise, leverages optimized code, less error-prone
**Cons**: Less transparent about what's happening internally
:::

## Approach 3: The Sorting Method {.smaller}

```{r}
# Method 3: Sort first, then take the top
find_max_sort <- function(data) {
  sorted_data <- data %>%
    arrange(desc(unemployment_rate))
  
  return(list(
    county = sorted_data$county[1],
    rate = sorted_data$unemployment_rate[1]
  ))
}

# Test it
result3 <- find_max_sort(unemployment_data)
cat("Highest unemployment:", result3$county, "at", 
    scales::percent(result3$rate, accuracy = 0.1))
```

::: {.fragment}
**Pros**: Very readable, gives us sorted data as a bonus
**Cons**: Does more work than necessary (sorts entire dataset)
:::

## Performance Comparison {.smaller}

Let's measure how these approaches perform with different data sizes:

```{r}
# Create datasets of different sizes
create_test_data <- function(n) {
  tibble(
    county = paste("County", 1:n),
    unemployment_rate = runif(n, 0.02, 0.15)
  )
}

# Test with 1000 rows
test_data <- create_test_data(1000)

# Compare performance
benchmark_results <- microbenchmark(
  simple = find_max_simple(test_data),
  builtin = find_max_builtin(test_data),
  sorting = find_max_sort(test_data),
  times = 100
)

print(benchmark_results)
```

## Visualizing Performance Differences {.smaller}

```{r}
#| echo: false
# Convert benchmark results to a more readable format
benchmark_summary <- benchmark_results %>%
  group_by(expr) %>%
  summarise(
    median_time_ms = median(time) / 1e6,  # Convert to milliseconds
    .groups = 'drop'
  ) %>%
  mutate(
    method = case_when(
      expr == "simple" ~ "Simple Loop",
      expr == "builtin" ~ "Built-in Function", 
      expr == "sorting" ~ "Sort Method"
    )
  )

ggplot(benchmark_summary, aes(x = reorder(method, median_time_ms), y = median_time_ms)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_text(aes(label = paste0(round(median_time_ms, 2), " ms")), 
            vjust = -0.5, size = 4) +
  labs(title = "Performance Comparison: Finding Maximum Value",
       subtitle = "Median time across 100 runs (1,000 counties)",
       x = "Method", y = "Time (milliseconds)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

::: {.fragment}
**Key insight**: The built-in function is dramatically faster! Even though sorting does extra work, R's optimized sorting is still faster than our simple loop.
:::

# Part III: Understanding Computational Complexity

## Big O Notation: The Language of Efficiency {.smaller}

::: {.fragment}
**Big O describes how runtime grows as input size increases**
:::

::: {layout-ncol=2}
### Common Complexities
- **O(1)**: Constant time
- **O(log n)**: Logarithmic
- **O(n)**: Linear  
- **O(n log n)**: Log-linear
- **O(n²)**: Quadratic
- **O(2ⁿ)**: Exponential

### Policy Translation
- **O(1)**: Checking eligibility rules
- **O(log n)**: Finding a record in sorted database
- **O(n)**: Processing all applications once
- **O(n²)**: Comparing every pair of cases
- **O(2ⁿ)**: Trying all possible combinations
:::

::: {.fragment}
**Critical insight**: The difference between O(n) and O(n²) can mean the difference between real-time and overnight processing.
:::

## Complexity in Action: Sorting Algorithms {.smaller}

Let's implement and compare two sorting approaches:

```{r}
# Bubble Sort: O(n²) - compares every pair
bubble_sort <- function(x) {
  n <- length(x)
  # Outer loop: we need n-1 passes to sort n elements
  # Each pass moves the largest unsorted element to its correct position
  for(i in 1:(n-1)) {
    # Inner loop: compare adjacent pairs from start to end of unsorted portion
    # We use (n-i) because after i passes, the last i elements are already sorted
    for(j in 1:(n-i)) {
      if(x[j] > x[j+1]) {
        # Swap elements
        temp <- x[j+1]
        x[j+1] <- x[j]
        x[j] <- temp
      }
    }
  }
  return(x)
}

# Test with small dataset
small_data <- c(64, 34, 25, 12, 22, 11, 90)
cat("Original:", paste(small_data, collapse = ", "), "\n")
cat("Sorted:  ", paste(bubble_sort(small_data), collapse = ", "), "\n")
```

## The Dramatic Impact of Complexity {.smaller}

Let's see how different complexities perform as data size grows:

```{r}
#| code-overflow: scroll
#| output-location: fragment
# Test sorting performance with different sizes
test_sorting_performance <- function(sizes) {
  results <- tibble()
  
  for(n in sizes) {
    # Create random data - ensure we have enough values to sample from
    max_val <- max(1000, n)  # Use at least 1000 or n, whichever is larger
    test_data <- sample(1:max_val, n)
    
    # Time bubble sort (O(n²))
    bubble_time <- system.time({
      bubble_sort(test_data)
    })["elapsed"]
    
    # Time R's built-in sort (O(n log n))
    builtin_time <- system.time({
      sort(test_data)
    })["elapsed"]
    
    results <- bind_rows(results, tibble(
      n = n,
      bubble_sort = bubble_time,
      builtin_sort = builtin_time
    ))
  }
  
  return(results)
}

# Test with increasing sizes (be careful not to make too large!)
performance_data <- test_sorting_performance(c(100, 300, 500, 800))
performance_data
```

## Visualizing the Complexity Explosion {.smaller}

```{r}
#| echo: false
# Reshape data for plotting
performance_long <- performance_data %>%
  pivot_longer(cols = c(bubble_sort, builtin_sort), 
               names_to = "algorithm", 
               values_to = "time_seconds") %>%
  mutate(
    algorithm = case_when(
      algorithm == "bubble_sort" ~ "Bubble Sort O(n²)",
      algorithm == "builtin_sort" ~ "Built-in Sort O(n log n)"
    )
  )

ggplot(performance_long, aes(x = n, y = time_seconds, color = algorithm)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = function(x) paste0(x, "s")) +
  scale_color_manual(values = c("Bubble Sort O(n²)" = "red", 
                               "Built-in Sort O(n log n)" = "blue")) +
  labs(title = "The Dramatic Impact of Algorithm Choice",
       subtitle = "How sorting time grows with data size",
       x = "Number of Items to Sort",
       y = "Time (seconds)",
       color = "Algorithm") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

::: {.fragment}
**Policy implication**: With 10,000 records, bubble sort might take hours while efficient sorting takes milliseconds!
:::

# Part IV: Policy-Relevant Algorithm Design

## Case Study: Emergency Resource Allocation {.smaller}

::: {.fragment}
**Scenario**: During a natural disaster, we need to allocate emergency supplies to affected areas.
:::

::: {.fragment}
**Constraints**:
- Limited supplies (trucks, medical kits, food)
- Different areas have different needs and priorities
- Time is critical
- Decisions must be auditable and fair
:::

::: {.fragment}
**Algorithm design questions**:
1. How do we prioritize areas?
2. How do we ensure fair distribution?
3. How do we handle real-time updates?
4. How do we make decisions transparent?
:::

## Algorithm 1: Simple Priority Queue {.smaller}

```{r}
# Create sample emergency data
set.seed(123)
emergency_data <- tibble(
  area_id = paste("Area", LETTERS[1:10]),
  population = sample(1000:50000, 10),
  severity_score = sample(1:10, 10, replace = TRUE),  # 1 = low, 10 = critical
  distance_km = sample(5:100, 10),
  supplies_needed = sample(10:500, 10)
) %>%
  # Calculate priority score (higher = more urgent)
  mutate(
    priority_score = (severity_score * population) / distance_km,
    priority_rank = rank(-priority_score)  # Negative for descending order
  )

head(emergency_data)
```

## Simple Allocation Algorithm {.smaller}

```{r}
# GREEDY ALLOCATION ALGORITHM
# Input: areas (tibble with priority_score, supplies_needed, etc.)
#        total_supplies (integer: total units available)
# Output: tibble with area_id, allocated, needed, priority_rank

allocate_supplies_simple <- function(areas, total_supplies) {
  
  # STEP 1: Sort areas by priority (highest priority first)
  # This creates our "queue" - we'll serve highest priority areas first
  sorted_areas <- areas %>%
    arrange(desc(priority_score))  # desc() = descending order
  
  # STEP 2: Initialize our tracking variables
  remaining_supplies <- total_supplies  # How many supplies we have left
  allocations <- tibble()              # Empty table to store results
  
  # STEP 3: Go through each area in priority order
  for(i in 1:nrow(sorted_areas)) {
    current_area <- sorted_areas[i, ]  # Get data for this area
    
    # STEP 4: Give this area as much as possible (but not more than needed)
    # Example: Area needs 500, we have 300 left → give 300
    # Example: Area needs 200, we have 300 left → give 200
    allocated_to_this_area <- min(current_area$supplies_needed, remaining_supplies)
    
    # STEP 5: Record what we gave to this area
    allocations <- bind_rows(allocations, tibble(
      area_id = current_area$area_id,           # Which area (character)
      allocated = allocated_to_this_area,       # How much we gave (integer)
      needed = current_area$supplies_needed,    # How much they needed (integer)
      priority_rank = current_area$priority_rank # Their rank (integer)
    ))
    
    # STEP 6: Update how many supplies we have left
    remaining_supplies <- remaining_supplies - allocated_to_this_area
    
    # STEP 7: If we're out of supplies, stop (no point continuing)
    if(remaining_supplies <= 0) break
  }
  
  # STEP 8: Add areas that got nothing (for complete reporting)
  # Find areas we never got to (ran out of supplies before reaching them)
  unallocated_areas <- sorted_areas %>%
    filter(!area_id %in% allocations$area_id) %>%  # Areas not in our results yet
    select(area_id, supplies_needed, priority_rank) %>%
    mutate(
      allocated = 0,                    # They got 0 supplies
      needed = supplies_needed          # Their original request
    ) %>%
    select(area_id, allocated, needed, priority_rank)  # Keep only the columns we want
  
  # STEP 9: Combine allocated and unallocated areas into final result
  final_allocations <- bind_rows(allocations, unallocated_areas)
  
  return(final_allocations)
}

# TEST: Run the algorithm with 1000 supply units
allocation_result <- allocate_supplies_simple(emergency_data, 1000)
allocation_result
```

## Evaluating Our Algorithm {.smaller}

```{r}
# Add allocation results back to original data for analysis
analysis_data <- allocation_result %>%
  mutate(
    percent_satisfied = allocated / needed,
    unmet_need = needed - allocated
  )

# Summary statistics
cat("Algorithm Performance Summary:\n")
cat("- Areas fully served:", sum(analysis_data$percent_satisfied >= 1), "out of", nrow(analysis_data), "\n")
cat("- Total supplies allocated:", sum(analysis_data$allocated, na.rm = TRUE), "out of 1000\n")
cat("- Average satisfaction rate:", scales::percent(mean(analysis_data$percent_satisfied, na.rm = TRUE)), "\n")
cat("- Highest priority area satisfaction:", 
    scales::percent(analysis_data$percent_satisfied[analysis_data$priority_rank == 1]), "\n")
```

## Algorithm Transparency and Auditability {.smaller}

```{r}
#| echo: false
# Create visualization of allocation decisions
ggplot(analysis_data, aes(x = reorder(area_id, -priority_rank))) +  # Negative to sort by priority
  geom_col(aes(y = needed), fill = "lightcoral", alpha = 0.7, width = 0.7) +
  geom_col(aes(y = allocated), fill = "steelblue", alpha = 0.9, width = 0.7) +
  geom_text(aes(y = pmax(needed, allocated) + 15, 
                label = paste("Rank", priority_rank)), 
            size = 3, angle = 45, hjust = 0) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +  # Add 15% buffer at top
  labs(title = "Emergency Supply Allocation Results",
       subtitle = "Red = Total Need, Blue = Actually Allocated (sorted by priority: Rank 1 = highest)",
       x = "Area (sorted by priority)", y = "Supply Units") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

::: {.fragment}
**Transparency benefits**: Stakeholders can see exactly how priorities were calculated and why certain areas received more resources.
:::

# Part V: Algorithm Design Principles for Policy

## The Policy Algorithm Design Framework {.smaller}

::: {.fragment}
**1. Define Success Clearly**
- What outcomes matter most? Often referred to as KPIs. 
- How do we measure fairness?
- What are acceptable trade-offs?
:::

::: {.fragment}
**2. Consider Stakeholder Needs**
- Who is affected by decisions?
- What information do they need?
- How can they contest decisions?
:::

::: {.fragment}
**3. Plan for Scale and Change**
- Will this work with 10x more data?
- How do we handle new requirements?
- Can we update the algorithm safely?
:::

::: {.fragment}
**4. Build in Accountability**
- Can we explain every decision?
- Can we audit the process?
- How do we detect and fix problems?
:::

## Common Algorithmic Patterns in Policy {.smaller}

::: {layout-ncol=2}
### Matching Problems
- **Examples**: School assignment, organ donation, job placement
- **Key considerations**: Stability, fairness, efficiency
- **Typical complexity**: O(n²) to O(n³)

### Optimization Problems
- **Examples**: Budget allocation, route planning, scheduling
- **Key considerations**: Multiple objectives, constraints
- **Typical complexity**: Often NP-hard, need approximations

### Classification Problems
- **Examples**: Fraud detection, risk assessment, eligibility
- **Key considerations**: False positive/negative rates, bias
- **Typical complexity**: O(n) for prediction, varies for training

### Search and Retrieval
- **Examples**: Finding records, matching cases, recommendations
- **Key considerations**: Accuracy, speed, privacy
- **Typical complexity**: O(log n) with good indexing
:::

## When Simple Algorithms Win {.smaller}

::: {.fragment}
**Policy contexts often favor simple approaches because:**

- **Transparency**: Stakeholders can understand the logic
- **Auditability**: Easier to verify correct implementation  
- **Maintainability**: Fewer bugs, easier to modify
- **Fairness**: Less likely to have hidden biases
- **Robustness**: Fail gracefully, predictable behavior
:::

::: {.fragment}
**Example**: Linear scoring models vs. complex ML for benefit eligibility
- Linear model: Each factor has clear weight, easy to explain
- Complex ML: Might be more accurate but creates "black box" decisions
:::

::: {.fragment}
::: callout-important
**Policy principle**: Choose the simplest algorithm that meets your accuracy and fairness requirements.
:::
:::

## Interactive Exercise: Algorithm Design Challenge {.smaller}

::: {.fragment}
**Scenario**: Design an algorithm for COVID-19 vaccine distribution during initial shortage.

**Constraints**:
- Limited doses (1000 per week)
- Multiple priority groups (healthcare workers, elderly, high-risk)
- Geographic equity concerns
- Need to track and report decisions

**Your task (5 minutes)**:
1. What factors should determine priority?
2. How would you handle ties?
3. What data would you need?
4. How would you ensure fairness?
:::

::: {.fragment}
**Discussion questions**:
- How do you balance individual risk vs. community benefit?
- What happens when priorities conflict?
- How do you maintain public trust in the process?
:::

# Part VI: Practical Implementation Considerations

## From Algorithm to Production System {.smaller}

::: {.fragment}
**Development considerations**:

- **Testing**: How do we verify correctness?
- **Performance**: Will it scale to real data sizes?
- **Monitoring**: How do we detect problems in production?
- **Updates**: How do we safely modify the algorithm?
:::

::: {.fragment}
**Policy-specific considerations**:

- **Documentation**: Can auditors understand the process?
- **Reproducibility**: Can we recreate any decision?
- **Compliance**: Does it meet legal requirements?
- **Public communication**: How do we explain it to citizens?
:::

## Algorithm Documentation Template {.smaller}

::: {.fragment}
**Essential documentation for policy algorithms:**

```markdown
## Algorithm: Emergency Supply Allocation

### Purpose
Allocate limited emergency supplies to affected areas during disasters

### Inputs
- Area data: population, severity, distance, needs
- Available supplies: total units available

### Logic
1. Calculate priority score: (severity × population) / distance
2. Sort areas by priority (highest first)  
3. Allocate supplies greedily until exhausted

### Assumptions
- Higher population areas need proportionally more supplies
- Closer areas can be served more efficiently
- Severity scores are accurate and comparable

### Limitations
- Does not consider supply chain constraints
- May not achieve optimal global allocation
- Assumes all supply units are equivalent

### Testing
- Verified with historical disaster data
- Stress tested with 10,000+ areas
- Edge cases: no supplies, no areas, invalid data
```
:::

## Looking Ahead: Advanced Topics {.smaller}

::: {.fragment}
**Next week and beyond, we'll explore:**

- **Machine learning algorithms**: When and how to use them in policy
- **Optimization techniques**: Finding the best solutions under constraints
- **Fairness-aware algorithms**: Ensuring equitable outcomes
- **Real-time systems**: Handling streaming data and immediate decisions
:::

::: {.fragment}
**Key preparation for ML topics:**

- Understanding of basic algorithmic complexity
- Appreciation for trade-offs between accuracy and interpretability
- Framework for evaluating algorithm quality in policy contexts
:::

## Questions?

::: {.fragment}
**Remember**: Good algorithmic thinking in policy isn't about finding the most sophisticated solution—it's about finding the right balance of accuracy, fairness, transparency, and efficiency for your specific context.
:::
