---
title: "Tree-Based Models for Complex Data"
subtitle: "Decision Trees, Random Forests, and Gradient Boosting"
author: "Rod Albuyeh"
date: "11/24/2025"
format:
  revealjs:
    theme: [default, simple]
    css: styles.css
    slide-number: true
    progress: true
    overview: true
    controls: true
    transition: fade
    auto-animate: true
    incremental: true
    code-overflow: wrap
    code-line-numbers: false
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r setup, include=TRUE, echo=FALSE, results=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 8, fig.height = 6)
library(tidyverse)
library(scales)
library(patchwork)
library(rpart)
library(rpart.plot)
library(randomForest)
library(ranger)
library(caret)
library(vip)
set.seed(42)
```

```{css echo=FALSE}
.reveal {
  font-size: 32px;
}

.reveal h2 {
  font-size: 1.3em;
}

.reveal p, .reveal li {
  font-size: 0.9em;
}

.reveal pre code {
  font-size: 0.85em;
}

.reveal .slides {
  text-align: center;
}

.reveal .slides section {
  text-align: left;
}
```

## Today's Learning Goals {.smaller}

::: {.fragment}
By the end of this session, you'll be able to:
:::

::: {.fragment}
- **Explain** how decision trees partition feature space through recursive splits
- **Distinguish** between individual trees, bagging, and boosting approaches
- **Apply** Random Forests and XGBoost to policy prediction problems
- **Interpret** variable importance and partial dependence in tree models
:::

::: {.fragment}
::: callout-note
**Key insight**: Tree-based models automatically capture non-linear relationships and interactions without feature engineering, making them powerful for complex policy data—but at the cost of some interpretability.
:::
:::

# Part I: From Linear Models to Trees

## Limitations of Linear Models {.smaller}

::: {.fragment}
**Last week we learned**: Regularization helps linear models handle many features
:::

::: {.fragment}
**But linear models still assume**:

- Relationships are linear (or manually transformed)
- Interactions must be explicitly specified
- Same relationship across all observations
:::

::: {.fragment}
**Policy example**: Predicting program enrollment
```r
# Linear model assumes same income effect for everyone
enrollment ~ income + age + children

# But reality might be:
# - Income matters more for families with children
# - Age effect differs by income level
# - Threshold effects (eligibility cutoffs)
```
:::

::: {.fragment}
::: callout-important
**Problem**: We'd need to manually specify all interactions and non-linearities. With many features, this becomes impractical!
:::
:::

## Enter Decision Trees: A Different Approach {.smaller}

::: {.fragment}
**Decision trees work by asking a series of yes/no questions:**
:::

::: {.fragment}
```
Q1: Does household have children? 
    → No: Predict 15% enrollment
    → Yes: Go to Q2
    
Q2: Is household income < $50K?
    → No: Predict 35% enrollment  
    → Yes: Go to Q3
    
Q3: Is distance to center < 5 miles?
    → No: Predict 40% enrollment
    → Yes: Predict 75% enrollment
```
:::

::: {.fragment}
**Key advantages**:

- **Automatic interactions**: Questions naturally create interactions
- **Non-linear relationships**: Different rules for different groups
- **Interpretable rules**: Easy to explain to stakeholders
- **No standardization needed**: Works with features on any scale
:::

## Visual Intuition: Partitioning Feature Space {.smaller}

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

# Create synthetic data with clear non-linear pattern
set.seed(123)
n <- 300
policy_data <- tibble(
  income = runif(n, 20000, 120000),
  distance_miles = runif(n, 0, 15)
) %>%
  mutate(
    prob = case_when(
      income < 50000 & distance_miles < 5 ~ 0.8,
      income < 50000 & distance_miles >= 5 ~ 0.4,
      income >= 50000 & distance_miles < 10 ~ 0.5,
      income >= 50000 & distance_miles >= 10 ~ 0.2,
      TRUE ~ 0.3
    ),
    enrolled = rbinom(n, 1, prob)
  ) %>%
  select(-prob)

# Create visualization with decision boundaries
ggplot(policy_data, aes(x = income, y = distance_miles, color = factor(enrolled))) +
  geom_point(size = 2, alpha = 0.6) +
  # Add tree partition lines
  geom_vline(xintercept = 50000, linetype = "dashed", color = "black", size = 1) +
  geom_segment(x = 20000, xend = 50000, y = 5, yend = 5, 
               linetype = "dashed", color = "black", size = 1) +
  geom_segment(x = 50000, xend = 120000, y = 10, yend = 10,
               linetype = "dashed", color = "black", size = 1) +
  # Add annotations for each partition
  annotate("text", x = 35000, y = 12, label = "Low Income\nFar (≥5 mi): 40%", size = 3.5, fontface = "bold") +
  annotate("text", x = 35000, y = 2.5, label = "Low Income\nNear (<5 mi): 80%", size = 3.5, fontface = "bold") +
  annotate("text", x = 85000, y = 12.5, label = "High Income\nFar (≥10 mi): 20%", size = 3.5, fontface = "bold") +
  annotate("text", x = 85000, y = 5, label = "High Income\nNear (<10 mi): 50%", size = 3.5, fontface = "bold") +
  scale_color_manual(values = c("0" = "coral", "1" = "steelblue"),
                     labels = c("Not Enrolled", "Enrolled"),
                     name = "Status") +
  scale_x_continuous(labels = dollar_format()) +
  labs(title = "How Decision Trees Partition Feature Space",
       subtitle = "Each rectangular region gets a different prediction",
       x = "Household Income", y = "Distance to Center (miles)") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

::: {.fragment}
**Key concept**: Trees create **rectangular partitions** in feature space, each with its own prediction.
:::

# Part II: How Decision Trees Work

## The Tree Building Algorithm {.smaller}

::: {.fragment}
**Core idea**: Recursively split data to minimize prediction error
:::

::: {.fragment}
**Algorithm (CART - Classification and Regression Trees)**:

1. **Start with all data** at the root node
2. **For each feature and each possible split point**:
   - Calculate prediction error if we split there
3. **Choose the split that reduces error the most**
4. **Create two child nodes** with the split data
5. **Repeat steps 2-4** for each child node
6. **Stop when**: 
   - Node is "pure" (all same outcome)
   - Reaches minimum sample size
   - Reaches maximum depth
:::

::: {.fragment}
::: callout-note
**Greedy algorithm**: At each step, makes the best split available *right now*, without looking ahead. Not guaranteed to find the globally optimal tree!
:::
:::

## Measuring Split Quality: Classification {.smaller}

::: {.fragment}
**For classification (e.g., enrolled vs not), we need a measure of "impurity"**
:::

::: {layout-ncol=2}
::: {.fragment}
### Gini Impurity
$$\text{Gini} = \sum_{k=1}^K p_k(1-p_k)$$

Where $p_k$ = proportion of class $k$

**Interpretation**:

- 0 = Pure (all same class)
- 0.5 = Maximum impurity (50/50 split)

**Most commonly used** in practice (default in most software)
:::

::: {.fragment}
### Entropy (Information Gain)
$$\text{Entropy} = -\sum_{k=1}^K p_k \log(p_k)$$

**Interpretation**:

- 0 = Pure
- Higher = More mixed

**From information theory**: Measures uncertainty
:::
:::

::: {.fragment}
**Example**: Node with 70% enrolled, 30% not enrolled

- Gini = 0.7(1-0.7) + 0.3(1-0.3) = 0.42
- Entropy = -0.7·log(0.7) - 0.3·log(0.3) ≈ 0.61
:::

## Split Selection Example {.smaller}

::: {.fragment}
**Scenario**: 100 families, 60 enrolled, 40 not enrolled
:::

```{r}
#| echo: false

split_comparison <- tibble(
  Split = c("Parent Node", "Split A: Income < $50K", "Split A: Income ≥ $50K", 
            "Split B: Has Children", "Split B: No Children"),
  N = c(100, 70, 30, 55, 45),
  Enrolled = c(60, 50, 10, 45, 15),
  `Not Enrolled` = c(40, 20, 20, 10, 30),
  Gini = c(0.48, 0.41, 0.44, 0.33, 0.44)
)

knitr::kable(split_comparison, 
             caption = "Comparing Two Possible Splits",
             digits = 2)
```

::: {.fragment}
**Weighted Gini after split**:

- **Split A**: (70/100)·0.41 + (30/100)·0.44 = 0.42
- **Split B**: (55/100)·0.33 + (45/100)·0.44 = 0.38 ← **Better!**

**Choose Split B** (has children?) because it reduces impurity more!
:::

## Building a Real Tree: Step-by-Step {.smaller}

```{r}
# Create sample enrollment data with clear patterns
set.seed(456)
n_families <- 200

enrollment_data <- tibble(
  family_id = 1:n_families,
  household_income = rnorm(n_families, 55000, 20000),
  has_children = sample(0:1, n_families, replace = TRUE, prob = c(0.5, 0.5)),
  distance_miles = rexp(n_families, rate = 0.2),
  previous_services = sample(0:1, n_families, replace = TRUE, prob = c(0.6, 0.4)),
  parent_age = rnorm(n_families, 32, 6)
) %>%
  mutate(
    household_income = pmax(15000, household_income),
    # True enrollment depends on these factors with strong interactions
    enroll_prob = plogis(-1.0 + 
                        -0.00003 * household_income + 
                        2.0 * has_children +
                        -0.25 * distance_miles + 
                        2.5 * previous_services +
                        1.0 * has_children * (household_income < 50000)),
    enrolled = rbinom(n_families, 1, enroll_prob)
  ) %>%
  select(-enroll_prob)

# Quick look
head(enrollment_data, 3)
cat("\nClass balance in full data:\n")
print(table(enrollment_data$enrolled))
cat("Proportion enrolled:", round(mean(enrollment_data$enrolled), 3), "\n")
```

## Fitting Our First Tree {.smaller}

```{r}
# Split data
set.seed(789)
train_idx <- createDataPartition(enrollment_data$enrolled, p = 0.7, list = FALSE)
train_data <- enrollment_data[train_idx, ]
test_data <- enrollment_data[-train_idx, ]

# Fit decision tree
tree_model <- rpart(
  enrolled ~ household_income + has_children + distance_miles + 
             previous_services + parent_age,
  data = train_data,
  method = "class",  # Classification tree
  control = rpart.control(
    cp = 0.005,       # Complexity parameter (like regularization) - lower allows more splits
    maxdepth = 5,     # Maximum tree depth
    minsplit = 15     # Minimum observations to attempt split
  )
)

cat("Tree structure created with", 
    sum(tree_model$frame$var == "<leaf>"), "leaf nodes\n")
```

## Visualizing the Tree {.smaller}

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7

# Create beautiful tree plot
rpart.plot(tree_model, 
          type = 4,  # Fancy style
          extra = 101,  # Show node probabilities and percentages
          fallen.leaves = TRUE,
          branch = 0.3,
          under = TRUE,
          box.palette = "auto",
          shadow.col = "gray",
          main = "Decision Tree for Program Enrollment",
          cex.main = 1.2,
          cex = 0.8)
```

::: {.fragment}
**Reading the tree**: 
- Each node shows: condition, predicted class, probability, % of data
- Follow branches left (yes) or right (no) based on conditions
:::

## Tree Predictions: A Walkthrough {.smaller}

::: {.fragment}
**Example family**: Income = $45K, has children, 3 miles away, no previous services
:::

::: {.fragment}
```
Root: Everyone starts here (60% enrolled overall)

├─ Previous services? NO → Go right
│  └─ Has children? YES → Go left  
│     └─ Distance < 5 miles? YES → Go left
│        └─ Prediction: 75% probability of enrollment → PREDICT: Enrolled
```
:::

::: {.fragment}
```{r}
# Make prediction for this family
new_family <- tibble(
  household_income = 45000,
  has_children = 1,
  distance_miles = 3,
  previous_services = 0,
  parent_age = 30
)

prediction <- predict(tree_model, new_family, type = "prob")
cat("Predicted probability of enrollment:", 
    round(prediction[1, "1"] * 100, 1), "%\n")
```
:::

::: {.fragment}
**Key insight**: Same prediction for ALL families in this leaf node!
:::

## Tree Performance {.smaller}

```{r}
# Predictions on test set
test_pred_prob <- predict(tree_model, test_data, type = "prob")[, "1"]
test_pred_class <- predict(tree_model, test_data, type = "class")

# Confusion matrix
conf_matrix <- table(Predicted = test_pred_class, Actual = test_data$enrolled)
print(conf_matrix)

# Calculate accuracy
accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)

# Compare to baseline (predict most common class)
baseline_pred <- rep(names(which.max(table(train_data$enrolled))), nrow(test_data))
baseline_accuracy <- mean(baseline_pred == test_data$enrolled)

cat("\nPerformance Comparison:\n")
cat("Baseline (always predict most common class):", round(baseline_accuracy * 100, 1), "%\n")
cat("Decision Tree:                              ", round(accuracy * 100, 1), "%\n")
cat("Improvement over baseline:                  ", round((accuracy - baseline_accuracy) * 100, 1), "percentage points\n")
```

::: {.fragment}
**Pretty good for a simple tree!** But we can do much better...
:::

# Part III: The Problem with Single Trees

## Overfitting and High Variance {.smaller}

::: {.fragment}
**Single trees have a critical weakness**: They overfit easily!
:::

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5

# Demonstrate tree instability with bootstrap samples
set.seed(111)
n_boots <- 5
boot_predictions <- matrix(0, nrow = nrow(test_data), ncol = n_boots)

for(i in 1:n_boots) {
  # Bootstrap sample
  boot_idx <- sample(1:nrow(train_data), replace = TRUE)
  boot_data <- train_data[boot_idx, ]
  
  # Fit tree
  boot_tree <- rpart(
    enrolled ~ household_income + has_children + distance_miles + 
               previous_services + parent_age,
    data = boot_data,
    method = "class",
    control = rpart.control(cp = 0.01, maxdepth = 4, minsplit = 20)
  )
  
  # Predict
  boot_predictions[, i] <- predict(boot_tree, test_data, type = "prob")[, "1"]
}

# Calculate prediction variance for each test observation
pred_variance <- apply(boot_predictions, 1, var)

# Visualize
variance_df <- tibble(
  observation = 1:nrow(test_data),
  avg_prediction = rowMeans(boot_predictions),
  variance = pred_variance,
  actual = test_data$enrolled
) %>%
  arrange(desc(variance))

ggplot(variance_df, aes(x = observation, y = avg_prediction)) +
  geom_point(aes(color = factor(actual)), size = 2) +
  geom_errorbar(aes(ymin = avg_prediction - sqrt(variance), 
                    ymax = avg_prediction + sqrt(variance)),
                alpha = 0.4, width = 0.5) +
  scale_color_manual(values = c("0" = "coral", "1" = "steelblue"),
                     labels = c("Not Enrolled", "Enrolled"),
                     name = "Actual Status") +
  labs(title = "Single Trees Are Unstable Across Bootstrap Samples",
       subtitle = "Error bars show prediction variance from 5 different bootstrap samples",
       x = "Test Observation (sorted by variance)", 
       y = "Average Predicted Probability") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

::: {.fragment}
**Problem**: Small changes in training data → dramatically different trees → different predictions
:::

## The Bias-Variance Problem Revisited {.smaller}

::: {.fragment}
**Decision trees trade bias for variance**:
:::

::: {layout-ncol=2}
::: {.fragment}
### Shallow Trees (High Bias)
- Simple partitions
- Smooth predictions
- **Underfit** the data
- Low variance, high bias
- Poor train AND test performance
:::

::: {.fragment}
### Deep Trees (High Variance)
- Complex partitions
- Jagged predictions  
- **Overfit** the data
- High variance, low bias
- Great train, poor test performance
:::
:::

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4

# Compare shallow vs deep trees
shallow_tree <- rpart(enrolled ~ ., data = train_data, method = "class",
                     control = rpart.control(maxdepth = 2, cp = 0.1))

deep_tree <- rpart(enrolled ~ ., data = train_data, method = "class",
                  control = rpart.control(maxdepth = 10, cp = 0.001, minsplit = 5))

# Predictions
shallow_train_acc <- mean(predict(shallow_tree, train_data, type = "class") == train_data$enrolled)
shallow_test_acc <- mean(predict(shallow_tree, test_data, type = "class") == test_data$enrolled)

deep_train_acc <- mean(predict(deep_tree, train_data, type = "class") == train_data$enrolled)
deep_test_acc <- mean(predict(deep_tree, test_data, type = "class") == test_data$enrolled)

comparison_df <- tibble(
  Model = c("Shallow Tree (maxdepth=2)", "Deep Tree (maxdepth=10)"),
  `Training Accuracy` = c(shallow_train_acc, deep_train_acc),
  `Test Accuracy` = c(shallow_test_acc, deep_test_acc),
  Gap = `Training Accuracy` - `Test Accuracy`
) %>%
  mutate(across(where(is.numeric), ~round(.x * 100, 1)))

knitr::kable(comparison_df, 
             caption = "The Overfitting Problem with Deep Trees")
```

::: {.fragment}
::: callout-important
**Solution**: Instead of building one perfect tree, build many imperfect trees and average them! This is called **ensemble learning**.
:::
:::

# Part IV: Random Forests - Wisdom of Crowds

## The Ensemble Idea {.smaller}

::: {.fragment}
**Key insight**: Many weak learners → one strong learner
:::

::: {.fragment}
**Analogy**: Which is more accurate?

1. **One expert's prediction** (single tree)
2. **Average of 500 experts' predictions** (random forest)

Usually #2, if experts have diverse perspectives!
:::

::: {.fragment}
**Random Forest recipe**:

1. **Bootstrap**: Create many random samples of training data (with replacement)
2. **Build trees**: Fit a decision tree to each bootstrap sample
3. **Randomize features**: At each split, only consider random subset of features
4. **Average predictions**: For classification, take majority vote; for regression, average
:::

::: {.fragment}
::: callout-note
**Why it works**: Individual trees overfit in different ways. Averaging cancels out the noise, keeps the signal!
:::
:::

## The Two Sources of Randomness {.smaller}

::: {.fragment}
**Random Forests use two types of randomness to create diversity**:
:::

::: {layout-ncol=2}
::: {.fragment}
### 1. Bootstrap Sampling (Bagging)
**Each tree sees different data**:
- Sample n observations with replacement
- ~63% of data in each bootstrap
- ~37% left out ("out-of-bag")

**Effect**: Trees see different patterns
:::

::: {.fragment}
### 2. Feature Randomness
**Each split considers random features**:
- Typically $\sqrt{p}$ features for classification
- Typically $p/3$ features for regression

**Effect**: Prevents strong features from dominating every tree
:::
:::

::: {.fragment}
```{r}
#| echo: false

tibble(
  `Problem Size` = c("5 features", "50 features", "500 features"),
  `Features per Split (Classification)` = c("√5 ≈ 2", "√50 ≈ 7", "√500 ≈ 22"),
  `Features per Split (Regression)` = c("5/3 ≈ 2", "50/3 ≈ 17", "500/3 ≈ 167")
) %>%
  knitr::kable(caption = "Typical mtry Values (features considered per split)")
```
:::

## Fitting a Random Forest {.smaller}

```{r}
# Fit Random Forest
rf_model <- randomForest(
  factor(enrolled) ~ household_income + has_children + distance_miles + 
                     previous_services + parent_age,
  data = train_data,
  ntree = 500,         # Number of trees
  mtry = 3,            # Features per split (try more than √5 for small data)
  importance = TRUE,   # Track feature importance
  nodesize = 5,        # Minimum node size (smaller for small dataset)
  maxnodes = 15        # Limit tree complexity
)

# Print results
print(rf_model)
```

::: {.fragment}
**OOB error**: Prediction error on out-of-bag samples (like built-in cross-validation!)
:::

## Random Forest Performance {.smaller}

```{r}
# Predictions on test set
rf_pred_prob <- predict(rf_model, test_data, type = "prob")[, "1"]
rf_pred_class <- predict(rf_model, test_data, type = "class")

# Confusion matrix
rf_conf_matrix <- table(Predicted = rf_pred_class, Actual = test_data$enrolled)
print(rf_conf_matrix)

# Accuracy comparison
rf_accuracy <- sum(diag(rf_conf_matrix)) / sum(rf_conf_matrix)

cat("\nAccuracy Comparison:\n")
cat("Baseline (most common class):  ", round(baseline_accuracy * 100, 1), "%\n")
cat("Single Tree:                   ", round(accuracy * 100, 1), "%\n")
cat("Random Forest:                 ", round(rf_accuracy * 100, 1), "%\n")
cat("RF Improvement over baseline:  ", round((rf_accuracy - baseline_accuracy) * 100, 1), "pp\n")
```

::: {.fragment}
**Note**: With small datasets (n=140 training), Random Forests may only marginally improve over a well-tuned single tree. The real power shows with larger datasets (n > 1000) where variance reduction is more substantial!
:::

## Variable Importance: What Matters Most? {.smaller}

::: {.fragment}
**One of the most useful features of Random Forests**: Variable importance scores
:::

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

# Extract and plot variable importance
importance_df <- as_tibble(randomForest::importance(rf_model), rownames = "Variable") %>%
  arrange(desc(MeanDecreaseGini))

ggplot(importance_df, aes(x = reorder(Variable, MeanDecreaseGini), 
                          y = MeanDecreaseGini)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  coord_flip() +
  labs(title = "Random Forest Variable Importance",
       subtitle = "Mean Decrease in Gini impurity when feature is used for splitting",
       x = "", y = "Importance Score") +
  theme_minimal()
```

::: {.fragment}
**Interpretation**: Household income and having children are the strongest predictors of enrollment.
:::

## How Variable Importance Works {.smaller}

::: {.fragment}
**Two main measures in Random Forests**:
:::

::: {layout-ncol=2}
::: {.fragment}
### Mean Decrease Gini (MDG)
**How it's calculated**:
- For each tree, for each feature
- Sum the decrease in Gini impurity when that feature is used to split
- Average across all trees

**Interpretation**:
- Higher = feature creates purer nodes
- Biased toward high-cardinality features

**When to use**: Quick screening, exploratory analysis
:::

::: {.fragment}
### Mean Decrease Accuracy (MDA)
**How it's calculated**:
- For each tree, for each feature
- Randomly permute that feature in OOB data
- Measure increase in prediction error
- Average across all trees

**Interpretation**:
- Higher = feature is more important for accurate predictions
- Less biased, more reliable

**When to use**: Final importance ranking, reporting to stakeholders
:::
:::

::: {.fragment}
::: callout-tip
**Policy recommendation**: Report **Mean Decrease Accuracy** for stakeholder-facing documents—it's more interpretable.
:::
:::

## Random Forest: Strengths and Limitations {.smaller}

::: {layout-ncol=2}
### Strengths ✓
::: {.fragment}
- **Excellent predictive accuracy** (often best off-the-shelf method)
- **Handles non-linear relationships** automatically
- **Captures interactions** without specification
- **No feature standardization** needed
- **Built-in variable importance**
- **Robust to outliers**
- **OOB error** (free validation estimate)
- **Few hyperparameters** to tune
:::

### Limitations ✗
::: {.fragment}
- **Black box**: Can't explain individual predictions easily
- **Computationally intensive** (500+ trees)
- **Memory hungry** (stores all trees)
- **Extrapolation poor** (can't predict outside training range)
- **Biased toward continuous features** and high-cardinality features
- **Less interpretable** than single trees or linear models
:::
:::

::: {.fragment}
::: callout-note
**Policy trade-off**: Random Forests sacrifice interpretability for accuracy. Use when prediction matters more than explanation!
:::
:::

# Part V: Gradient Boosting - Sequential Learning

## A Different Ensemble Strategy {.smaller}

::: {.fragment}
**Random Forest**: Build many trees **in parallel**, each independent
:::

::: {.fragment}
**Gradient Boosting**: Build trees **sequentially**, each learning from previous mistakes
:::

::: {.fragment}
**The boosting algorithm**:

1. **Start simple**: Fit shallow tree (just a few splits)
2. **Find mistakes**: Calculate residuals (actual - predicted)
3. **Learn from mistakes**: Fit new tree to predict the residuals
4. **Add correction**: Update predictions = old predictions + (learning rate × new tree)
5. **Repeat**: Keep adding trees that correct previous errors
6. **Final prediction**: Sum of all tree contributions
:::

::: {.fragment}
::: callout-note
**Key insight**: Instead of averaging independent trees (Random Forest), boosting creates a **committee** where each member focuses on what previous members got wrong!
:::
:::

## Boosting Intuition: A Visual {.smaller}

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 6

# Create simple visualization of boosting process
set.seed(999)
x <- seq(0, 10, length.out = 100)
true_y <- sin(x) + 0.3 * x
noisy_y <- true_y + rnorm(100, 0, 0.5)

# Simulate boosting stages
stage1 <- rep(mean(noisy_y), 100)  # Start with mean
residuals1 <- noisy_y - stage1
stage2 <- stage1 + 0.3 * (residuals1)  # Add first correction
residuals2 <- noisy_y - stage2  
stage3 <- stage2 + 0.3 * (residuals2)  # Add second correction

boosting_df <- tibble(
  x = rep(x, 4),
  y = c(noisy_y, stage1, stage2, stage3),
  stage = factor(rep(c("Data", "Stage 1 (mean)", "Stage 2 (+tree 1)", "Stage 3 (+tree 2)"),
                     each = 100),
                 levels = c("Data", "Stage 1 (mean)", "Stage 2 (+tree 1)", "Stage 3 (+tree 2)"))
)

ggplot(boosting_df, aes(x = x, y = y)) +
  geom_point(data = filter(boosting_df, stage == "Data"), 
             alpha = 0.3, color = "gray40") +
  geom_line(data = filter(boosting_df, stage != "Data"),
            aes(color = stage), size = 1.2) +
  scale_color_viridis_d(begin = 0.3, end = 0.9) +
  facet_wrap(~stage, ncol = 4) +
  labs(title = "How Gradient Boosting Learns Sequentially",
       subtitle = "Each stage adds a correction to reduce residual errors",
       x = "Feature", y = "Outcome") +
  theme_minimal() +
  theme(legend.position = "none")
```

::: {.fragment}
**Notice**: Predictions get progressively better as we add more trees!
:::

## Key Hyperparameters in Gradient Boosting {.smaller}

::: {.fragment}
**Boosting has more knobs to turn than Random Forests**:
:::

::: {layout-ncol=2}
::: {.fragment}
### Learning Rate (η)
- **What**: How much each tree contributes
- **Typical**: 0.01 to 0.3
- **Small η**: Needs more trees, but better generalization
- **Large η**: Fewer trees, but risks overfitting

**Analogy**: Step size when walking toward solution
:::

::: {.fragment}
### Number of Trees
- **What**: How many sequential corrections
- **Typical**: 100 to 10,000+
- **Too few**: Underfits
- **Too many**: Can overfit (but slowly with small η)

**Analogy**: How many steps to take
:::

::: {.fragment}
### Tree Depth (max_depth)
- **What**: Complexity of each individual tree
- **Typical**: 3 to 8
- **Shallow**: Simple corrections, needs more trees
- **Deep**: Complex corrections, risks overfitting

**Analogy**: How complex each step is
:::

::: {.fragment}
### Subsample
- **What**: % of data used per tree
- **Typical**: 0.5 to 1.0
- **Adds randomness** (like Random Forest)
- **Stochastic gradient boosting**

**Analogy**: Practice on subset before full game
:::
:::

## XGBoost: Extreme Gradient Boosting {.smaller}

::: {.fragment}
**XGBoost** = Most popular gradient boosting implementation
:::

::: {.fragment}
**Why XGBoost dominates Kaggle competitions**:

- **Regularization**: Built-in L1/L2 penalties (like LASSO/Ridge)
- **Speed**: Highly optimized C++ implementation
- **Handles missing data**: Built-in imputation strategy
- **Parallel processing**: Distributes tree building
- **Custom loss functions**: Optimize any objective
- **Early stopping**: Automatically stops when validation error stops improving
:::

::: {.fragment}
```{r}
#| eval: false

# XGBoost in R (using caret interface)
library(xgboost)

xgb_model <- train(
  factor(enrolled) ~ .,
  data = train_data,
  method = "xgbTree",
  trControl = trainControl(method = "cv", number = 5),
  tuneGrid = expand.grid(
    nrounds = c(100, 200, 500),          # Number of trees
    max_depth = c(3, 5, 7),              # Tree depth
    eta = c(0.01, 0.1, 0.3),             # Learning rate
    gamma = 0,                            # Minimum split loss
    colsample_bytree = 0.8,              # Feature sampling
    min_child_weight = 1,                # Minimum sum of weights
    subsample = 0.8                      # Row sampling
  )
)
```
:::

## Random Forest vs Gradient Boosting {.smaller}

```{r}
#| echo: false

comparison_table <- tibble(
  Aspect = c("Building Strategy", "Tree Depth", "Bias-Variance Trade-off", 
             "Training Speed", "Overfitting Risk", "Hyperparameter Tuning",
             "Interpretability", "When to Use"),
  `Random Forest` = c(
    "Parallel (independent)",
    "Deep trees",
    "Reduces variance",
    "Faster (parallelizable)",
    "Low (averaging helps)",
    "Easy (few parameters)",
    "Moderate (variable importance)",
    "General purpose, robust"
  ),
  `Gradient Boosting` = c(
    "Sequential (learn from mistakes)",
    "Shallow trees",
    "Reduces bias",
    "Slower (sequential)",
    "Higher (needs careful tuning)",
    "Complex (many parameters)",
    "Lower (complex interactions)",
    "Competitions, maximum accuracy"
  )
)

knitr::kable(comparison_table, caption = "Comparing Ensemble Strategies")
```

::: {.fragment}
::: callout-tip
**Practical advice**: Start with **Random Forest** (easier, more robust). Try **XGBoost** if you need maximum accuracy and have time to tune.
:::
:::

# Part VI: Model Interpretation for Policy

## The Interpretability Challenge {.smaller}

::: {.fragment}
**Problem**: Random Forests and XGBoost are "black boxes"
:::

::: {.fragment}
**Policy stakeholders ask**:

- "Why did this family get flagged?"
- "What should we change to improve enrollment?"
- "Can you explain this in plain English?"
:::

::: {.fragment}
**Three approaches to interpretability**:

1. **Global importance**: Which features matter overall?
2. **Partial dependence**: How does each feature affect predictions?
3. **SHAP values**: Why did this specific prediction happen?
:::

::: {.fragment}
::: callout-important
**Policy reality**: You often need to sacrifice some accuracy for interpretability. A slightly worse model you can explain is better than a perfect model you can't defend!
:::
:::

## Global Variable Importance {.smaller}

::: {.fragment}
**We've already seen this for Random Forest**:
:::

```{r}
#| fig-width: 9
#| fig-height: 4

# Use vip package for better visualization
library(vip)

vip(rf_model, num_features = 5, geom = "col", 
    aesthetics = list(fill = "steelblue", alpha = 0.8)) +
  labs(title = "Top 5 Most Important Features",
       subtitle = "Random Forest Variable Importance (Mean Decrease Accuracy)") +
  theme_minimal()
```

::: {.fragment}
**Stakeholder message**: "The strongest predictors of enrollment are previous service use and having children in the household."
:::

## Partial Dependence Plots (PDP) {.smaller}

::: {.fragment}
**Question**: How does enrollment probability change with income, *holding everything else constant*?
:::

::: {.fragment}
**How PDPs work**:

1. Pick a feature (e.g., income)
2. For each value of income (e.g., $30K, $40K, $50K...):
   - Set everyone's income to that value
   - Keep all other features at their original values
   - Make predictions for everyone
   - Average the predictions
3. Plot average prediction vs feature value
:::

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5

# Create manual PDP for income
income_seq <- seq(min(train_data$household_income), 
                  max(train_data$household_income), 
                  length.out = 50)

pdp_predictions <- map_dbl(income_seq, function(inc) {
  temp_data <- train_data
  temp_data$household_income <- inc
  preds <- predict(rf_model, temp_data, type = "prob")[, "1"]
  mean(preds)
})

pdp_df <- tibble(
  household_income = income_seq,
  enrollment_prob = pdp_predictions
)

ggplot(pdp_df, aes(x = household_income, y = enrollment_prob)) +
  geom_line(color = "steelblue", size = 1.5) +
  geom_ribbon(aes(ymin = enrollment_prob - 0.02, 
                  ymax = enrollment_prob + 0.02),
              alpha = 0.2, fill = "steelblue") +
  scale_x_continuous(labels = dollar_format()) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Partial Dependence: Income Effect on Enrollment",
       subtitle = "How predicted enrollment probability changes with income (holding other features constant)",
       x = "Household Income", 
       y = "Predicted Enrollment Probability") +
  theme_minimal()
```

::: {.fragment}
**Insight**: Enrollment probability decreases as income increases—wealthier families less likely to enroll.
:::

## Interaction Effects: 2D PDPs {.smaller}

::: {.fragment}
**Trees automatically capture interactions**. We can visualize them:
:::

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

# Create 2D PDP for income × children interaction
income_grid <- seq(min(train_data$household_income), 
                   max(train_data$household_income), length.out = 30)
children_grid <- c(0, 1)

pdp_2d <- expand.grid(
  household_income = income_grid,
  has_children = children_grid
) %>%
  mutate(
    enrollment_prob = map2_dbl(household_income, has_children, function(inc, kids) {
      temp_data <- train_data
      temp_data$household_income <- inc
      temp_data$has_children <- kids
      mean(predict(rf_model, temp_data, type = "prob")[, "1"])
    })
  )

ggplot(pdp_2d, aes(x = household_income, y = enrollment_prob, 
                   color = factor(has_children), 
                   linetype = factor(has_children))) +
  geom_line(size = 1.5) +
  scale_color_manual(values = c("0" = "coral", "1" = "steelblue"),
                     labels = c("No Children", "Has Children"),
                     name = "") +
  scale_linetype_manual(values = c("0" = "dashed", "1" = "solid"),
                        labels = c("No Children", "Has Children"),
                        name = "") +
  scale_x_continuous(labels = dollar_format()) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  labs(title = "Interaction Effect: Income × Children",
       subtitle = "Income matters more for families WITH children",
       x = "Household Income",
       y = "Predicted Enrollment Probability") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

::: {.fragment}
**Key finding**: The income effect is stronger for families with children—an interaction the model discovered automatically!
:::

## Practical Interpretation Guidelines {.smaller}

::: {.fragment}
**For policy stakeholders, provide**:
:::

::: {.fragment}
1. **Variable importance**: Top 5-10 drivers (bar chart)
2. **Partial dependence**: 2-3 most important features (line charts)
3. **Performance metrics**: Accuracy, precision, recall (confusion matrix)
4. **Model card**: Summary document with use cases, limitations, bias audit
:::

::: {.fragment}
**What NOT to do**:

- Don't show tree diagrams for Random Forests (500 trees is overwhelming!)
- Don't report coefficients (trees don't have them!)
- Don't claim causality (still just correlations!)
- Don't hide model limitations
:::

::: {.fragment}
::: callout-note
**Communication principle**: Focus on *what* the model learned, not *how* it learned it. Stakeholders care about insights, not algorithms!
:::
:::

# Part VII: Practical Guidelines and Best Practices

## Choosing the Right Tree Method {.smaller}

```{r}
#| echo: false

decision_guide <- tibble(
  Scenario = c(
    "Need to explain to non-technical audience",
    "Maximum predictive accuracy required",
    "Limited computational resources",
    "Many features, correlated",
    "Many categorical features (high cardinality)",
    "Very large datasets (millions of rows)",
    "Non-linear relationships, unknown interactions",
    "Regulatory/legal scrutiny expected",
    "Lots of missing data",
    "Need GPU acceleration"
  ),
  Recommendation = c(
    "Single Decision Tree (shallow)",
    "XGBoost or LightGBM (with tuning)",
    "Single Tree or Random Forest (small)",
    "Random Forest or Elastic Net",
    "CatBoost (native categorical handling)",
    "LightGBM (fastest for big data)",
    "Random Forest or XGBoost",
    "Interpretable model (Tree, Linear) + SHAP",
    "XGBoost or CatBoost (native missing support)",
    "LightGBM or XGBoost (GPU support)"
  )
)

knitr::kable(decision_guide, caption = "Decision Guide: Which Tree Method?")
```

## Implementation Workflow {.smaller}

::: {.fragment}
**Best practices for tree-based models in policy**:
:::

::: {.fragment}
**1. Start simple, add complexity**
```r
# Step 1: Fit baseline single tree
single_tree <- rpart(outcome ~ ., data = train, maxdepth = 3)

# Step 2: Try Random Forest with defaults  
rf_default <- randomForest(outcome ~ ., data = train)

# Step 3: Tune Random Forest
rf_tuned <- train(outcome ~ ., data = train, method = "rf",
                 tuneGrid = expand.grid(mtry = c(2, 5, 10)))

# Step 4: If needed, try XGBoost
xgb_model <- train(outcome ~ ., data = train, method = "xgbTree")
```
:::

::: {.fragment}
**2. Always compare to simpler baselines**
- Logistic regression
- Elastic net
- Single decision tree
:::

::: {.fragment}
**3. Validate thoroughly**
- Cross-validation during training
- Hold-out test set for final evaluation
- Check performance across subgroups (fairness)
:::

## Common Pitfalls to Avoid {.smaller}

::: {layout-ncol=2}
### Technical Mistakes
::: {.fragment}
1. **Not tuning hyperparameters**
   - Default settings rarely optimal
   
2. **Overfitting with deep trees**
   - Use cross-validation to pick depth
   
3. **Ignoring class imbalance**
   - Use class weights or SMOTE
   
4. **Not checking feature importance**
   - May reveal data leakage
   
5. **Extrapolation beyond training range**
   - Trees can't extrapolate!
:::

### Interpretation Mistakes
::: {.fragment}
6. **Confusing prediction with causation**
   - Trees find correlations, not causes
   
7. **Over-interpreting variable importance**
   - Rankings can be unstable
   
8. **Ignoring model uncertainty**
   - Provide confidence intervals
   
9. **Forgetting about fairness**
   - Check disparate impact across groups
   
10. **Not documenting model choices**
    - Keep record of why you picked this model
:::
:::

## Fairness and Bias Considerations {.smaller}

::: {.fragment}
**Tree models can encode and amplify bias**:
:::

::: {.fragment}
**Example**: Historical lending data

- Past discrimination → Biased training data
- Model learns: "Deny loans to ZIP codes X, Y, Z"
- These ZIP codes happen to be predominantly minority
- **Result**: Model perpetuates historical discrimination
:::

::: {.fragment}
**Mitigation strategies**:

1. **Audit feature importance**: Do sensitive features (race, gender) matter directly or indirectly?
2. **Check subgroup performance**: Does accuracy differ by protected groups?
3. **Remove proxy features**: ZIP code, names, etc. that correlate with protected attributes
4. **Fairness constraints**: Train with fairness-aware objectives
5. **Human review**: Don't fully automate high-stakes decisions
:::

::: {.fragment}
::: callout-important
**Policy principle**: Higher accuracy doesn't excuse discrimination. Fairness is a constraint, not an optimization target!
:::
:::

## When NOT to Use Tree Models {.smaller}

::: {.fragment}
**Trees aren't always the answer**:
:::

::: {layout-ncol=2}
::: {.fragment}
### Better Alternatives Exist
**Use linear models when**:
- Relationships are roughly linear
- Need coefficient interpretation
- Want statistical inference (p-values)
- Have few features (<20)
- Need to extrapolate

**Example**: Credit risk assessment
:::

::: {.fragment}
### Other Considerations
**Don't use trees when**:
- Full transparency required by law
- Stakeholders demand explainability
- Computational resources very limited
- Real-time prediction critical (<1ms)
- Training data very small (n < 100)

**Example**: Medical diagnosis requiring FDA approval
:::
:::

::: {.fragment}
::: callout-tip
**Decision rule**: If linear/logistic regression achieves 90% of tree performance, use the simpler model for policy work!
:::
:::

# Part VIII: Looking Forward

## Trees in Modern ML Pipelines {.smaller}

::: {.fragment}
**Tree-based models are often just one component**:
:::

::: {.fragment}
**Typical policy ML pipeline**:

1. **Feature engineering** (manual + automated)
2. **Try multiple model families**:
   - Elastic Net (interpretable baseline)
   - Random Forest (robust default)
   - XGBoost (maximum accuracy)
3. **Ensemble different models** (stack them)
4. **Select based on trade-offs** (accuracy vs interpretability vs computational cost)
5. **Deploy with monitoring** (detect drift, retrain)
:::

::: {.fragment}
**Real-world example**: Fraud detection system
- **Layer 1**: Rules-based filters (transparent, fast)
- **Layer 2**: Logistic regression (interpretable)
- **Layer 3**: Random Forest (final check, high accuracy)
- **Human review**: Uncertain cases
:::

## Emerging Trends {.smaller}

::: {.fragment}
**What's next for tree-based models**:
:::

::: {layout-ncol=2}
::: {.fragment}
### Algorithmic Advances
- **LightGBM**: Even faster than XGBoost
- **CatBoost**: Better handling of categorical features
- **NGBoost**: Probabilistic predictions with uncertainty
- **AutoML**: Automated end to end modeling pipeline
:::

::: {.fragment}
### Interpretability Tools
- **SHAP values**: Unified framework for model explanations
- **LIME**: Local interpretable approximations
- **TreeSHAP**: Fast SHAP for trees
- **Counterfactual explanations**: "What if" analysis
:::
:::

::: {.fragment}
::: callout-note
**For next week**: We'll explore unsupervised learning—clustering, PCA, and dimension reduction techniques for discovering patterns without labels!
:::
:::

## Key Takeaways {.smaller}

::: {.fragment}
1. **Decision trees partition feature space** through recursive binary splits, creating interpretable rules
:::

::: {.fragment}
2. **Single trees overfit easily** (high variance), making them unstable across different samples
:::

::: {.fragment}
3. **Random Forests reduce variance** by averaging many diverse trees (bootstrap + feature randomness)
:::

::: {.fragment}
4. **Gradient Boosting reduces bias** by sequentially learning from mistakes (XGBoost is the standard)
:::

::: {.fragment}
5. **Variable importance shows what matters**, partial dependence shows how features affect predictions
:::

::: {.fragment}
6. **Trees handle non-linearity and interactions automatically**, but sacrifice interpretability
:::

::: {.fragment}
7. **Start with Random Forest** (robust, easy), try XGBoost if you need maximum accuracy
:::

::: {.fragment}
8. **Always check for bias** and validate fairness across protected groups
:::

::: {.fragment}
**Looking ahead**:

- **Week 10** (Dec 1): AI-augmented policy analysis, LLM APIs
- **Final Exam** (Dec 3): In-class, cumulative
- **Assignment 3** due Sat Dec 6 (final project report)
:::

## Course Evaluations: A Collective Action Problem {.smaller}

::: {.fragment}
**Challenge**: Course evaluations have low response rates, yet they're critical for improving teaching
:::

::: {.fragment}
**Incentive structure**: If ≥90% of the class completes evaluations by [deadline], **everyone** gets +0.5% added to their final grade
:::

::: {.fragment}
**"Wait, the course is curved—does this help me?"**

Yes! Here's why:
:::

::: {layout-ncol=2}
::: {.fragment}
### How It Works
- +0.5% added to your **raw score** before curving
- Think of it as **catastrophe insurance**
:::

::: {.fragment}
### The Policy Design
- **Positive externality**: Your response helps everyone
- **Coordination game**: Need 90% cooperation (Nash equilibrium at high participation)
- **Low individual cost**: ~5 minutes
- **Collective benefit**: Buffer against grade anxiety + better course feedback
:::
:::

::: {.fragment}
::: callout-tip
**Game theory insight**: Not responding costs **both you AND your classmates** the 0.5% buffer. Your self-interest aligns with the group's interest—classic coordination game where cooperation dominates!
:::
:::

## Questions?

::: {.fragment}
**Remember**: Tree-based models are among the most powerful tools in applied machine learning. They automatically discover non-linear relationships and interactions, making them ideal for complex policy data—but always balance accuracy with interpretability and fairness!
:::

---

