---
title: "Unsupervised Learning and Dimensionality Reduction"
subtitle: "Finding Hidden Patterns Without Labels"
author: "Rod Albuyeh"
date: "11/26/2025"
format:
  revealjs:
    theme: [default, simple]
    css: styles.css
    slide-number: true
    progress: true
    overview: true
    controls: true
    transition: fade
    auto-animate: true
    incremental: true
    code-overflow: wrap
    code-line-numbers: false
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r setup, include=TRUE, echo=FALSE, results=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 8, fig.height = 6)
library(tidyverse)
library(scales)
library(patchwork)
library(cluster)
library(factoextra)
library(ggfortify)
library(GGally)
set.seed(42)
```

```{css echo=FALSE}
.reveal {
  font-size: 32px;
}

.reveal h2 {
  font-size: 1.3em;
}

.reveal p, .reveal li {
  font-size: 0.9em;
}

.reveal pre code {
  font-size: 0.85em;
}

.reveal .slides {
  text-align: center;
}

.reveal .slides section {
  text-align: left;
}
```

## Today's Learning Goals {.smaller}

::: {.fragment}
By the end of this session, you'll be able to:
:::

::: {.fragment}
- **Distinguish** between supervised and unsupervised learning approaches
- **Apply** k-means and hierarchical clustering to discover groups in policy data
- **Interpret** principal component analysis (PCA) for dimensionality reduction
- **Evaluate** clustering quality using appropriate metrics
:::

::: {.fragment}
::: callout-note
**Key insight**: Unsupervised learning discovers hidden patterns and structure in data without labeled outcomes—essential for exploration, segmentation, and understanding complex policy landscapes before building predictive models.
:::
:::

# Part I: What Is Unsupervised Learning?

## Supervised vs Unsupervised: The Fundamental Difference {.smaller}

::: {.fragment}
**Everything we've learned so far has been supervised learning**:
:::

::: {layout-ncol=2}
::: {.fragment}
### Supervised Learning
**We have labels**: $(y_i, \mathbf{x}_i)$

**Examples**:
- Predict program enrollment (yes/no)
- Predict poverty rate (continuous)
- Classify fraud (fraud/legitimate)

**Goal**: Learn $\hat{f}(\mathbf{x})$ to predict $y$

**Evaluation**: Compare predictions to actual labels
:::

::: {.fragment}
### Unsupervised Learning
**No labels**: just $(\mathbf{x}_i)$

**Examples**:
- Group similar counties
- Find policy archetypes
- Reduce 100 survey questions to 5 themes

**Goal**: Discover structure, patterns, groups

**Evaluation**: No "ground truth"—must use domain knowledge + internal metrics
:::
:::

::: {.fragment}
::: callout-important
**Critical difference**: Unsupervised learning is exploratory. There's no single "right answer"—just useful patterns that help us understand the data!
:::
:::

## Why Unsupervised Learning Matters in Policy {.smaller}

::: {.fragment}
**Real-world policy scenarios where unsupervised learning shines**:
:::

::: {.fragment}
**1. Segmentation for Targeted Interventions**
- **Problem**: 3,000 counties, limited resources, which ones are similar?
- **Unsupervised approach**: Cluster counties by demographics/economics/health
- **Outcome**: "Counties in Cluster 3 need job training, Cluster 5 needs healthcare access"
:::

::: {.fragment}
**2. Simplifying Complex Survey Data**
- **Problem**: 200 questions measuring "community well-being"
- **Unsupervised approach**: PCA to find 5-10 underlying dimensions
- **Outcome**: "Questions reduce to: Economic Security, Health Access, Social Capital, Education, Safety"
:::

::: {.fragment}
**3. Discovering Policy Typologies**
- **Problem**: What types of cities exist based on policy choices?
- **Unsupervised approach**: Cluster cities by their policy portfolios
- **Outcome**: Identify "Progressive Urban," "Conservative Suburban," "Mixed Rural" archetypes
:::

::: {.fragment}
**4. Anomaly Detection**
- **Problem**: Find unusual patterns in government spending
- **Unsupervised approach**: Identify outliers that don't fit normal clusters
- **Outcome**: Flag potential fraud or data entry errors
:::

## The Two Main Families {.smaller}

::: {.fragment}
**Unsupervised learning splits into two major approaches**:
:::

::: {layout-ncol=2}
::: {.fragment}
### Clustering
**Goal**: Group similar observations together

**Questions answered**:
- Which observations are alike?
- How many distinct groups exist?
- What characterizes each group?

**Methods**:
- k-means clustering
- Hierarchical clustering
- DBSCAN, Gaussian Mixture Models

**Output**: Group assignments (e.g., "County 1 is in Cluster 3")
:::

::: {.fragment}
### Dimensionality Reduction
**Goal**: Simplify many variables into fewer dimensions

**Questions answered**:
- What underlying factors drive variation?
- Can we visualize high-dimensional data?
- Which variables are redundant?

**Methods**:
- Principal Component Analysis (PCA)
- Autoencoders
- Factor Analysis

**Output**: New variables (e.g., "PC1 captures 45% of variation")
:::
:::

::: {.fragment}
**Today**: We'll dive deep into k-means, hierarchical clustering, and PCA—the foundational techniques.
:::

# Part II: Clustering - Finding Natural Groups

## The Clustering Problem {.smaller}

::: {.fragment}
**Setup**: You have n observations with p features, but no labels. How do you find groups?
:::

::: {.fragment}
**Intuition**: Observations in the same group should be more similar to each other than to observations in other groups
:::

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

# Create synthetic policy data with clear clusters
set.seed(123)
n_counties <- 150

# Create three separate cluster groups
cluster1 <- tibble(
  median_income = rnorm(50, 80000, 10000),
  unemployment = rnorm(50, 3.5, 0.8),
  cluster = "1"
)

cluster2 <- tibble(
  median_income = rnorm(50, 35000, 8000),
  unemployment = rnorm(50, 8.5, 1.2),
  cluster = "2"
)

cluster3 <- tibble(
  median_income = rnorm(50, 60000, 10000),
  unemployment = rnorm(50, 5.5, 1.0),
  cluster = "3"
)

# Combine and create labels
cluster_data <- bind_rows(cluster1, cluster2, cluster3) %>%
  mutate(
    cluster_label = factor(case_when(
      cluster == "1" ~ "Urban\n(High Income, Low Unemployment)",
      cluster == "2" ~ "Rural\n(Low Income, High Unemployment)",
      cluster == "3" ~ "Suburban\n(Medium Income, Medium Unemployment)"
    ))
  )

ggplot(cluster_data, aes(x = median_income, y = unemployment, color = cluster_label)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_x_continuous(labels = dollar_format()) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  scale_color_viridis_d(name = "True Groups\n(unknown in practice)") +
  labs(title = "County Economic Profiles: Natural Groupings",
       subtitle = "Can we discover these 3 groups automatically?",
       x = "Median Household Income",
       y = "Unemployment Rate") +
  theme_minimal() +
  theme(legend.position = "right",
        legend.text = element_text(size = 9))
```

::: {.fragment}
**The challenge**: In real data, we don't know the true groups. We need algorithms to discover them!
:::

## k-Means Clustering: The Algorithm {.smaller}

::: {.fragment}
**k-means** is the most popular clustering algorithm. It's simple, fast, and often works well.
:::

::: {.fragment}
**The algorithm (Lloyd's algorithm)**:

1. **Choose k**: Decide how many clusters you want
2. **Initialize**: Randomly place k "centroids" (cluster centers)
3. **Assign**: Assign each observation to nearest centroid
4. **Update**: Move each centroid to the mean of its assigned points
5. **Repeat steps 3-4** until centroids stop moving (convergence)
:::

::: {.fragment}
**Objective**: Minimize within-cluster sum of squares (WCSS)

$$\text{WCSS} = \sum_{k=1}^K \sum_{i \in C_k} \|\mathbf{x}_i - \boldsymbol{\mu}_k\|^2$$

Where:
- $C_k$ = set of observations assigned to cluster $k$
- $i \in C_k$ = "for all observations $i$ in cluster $k$"
- $\boldsymbol{\mu}_k$ = centroid (mean) of cluster $k$
- $\|\mathbf{x}_i - \boldsymbol{\mu}_k\|^2$ = squared distance from observation $i$ to its cluster center
:::

::: {.fragment}
::: callout-note
**Key insight**: k-means finds clusters by making observations within each cluster as close to their center as possible!
:::
:::

## k-Means: Visual Walkthrough {.smaller}

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7

# Demonstrate k-means iterations with deliberately poor initialization
set.seed(456)
sample_data <- cluster_data %>%
  select(median_income, unemployment) %>%
  sample_n(90) %>%
  as.matrix()

# Start with TERRIBLE initial centers (far from the actual clusters)
# These are in completely wrong regions of the space
init_centers <- matrix(c(
  30000, 4.0,    # Center 1: far left, low unemployment (wrong!)
  55000, 11.0,   # Center 2: middle, very high unemployment (wrong!)
  85000, 7.0     # Center 3: far right, high unemployment (wrong!)
), ncol = 2, byrow = TRUE)

# Manually compute k-means iterations to show the actual process
# Iteration 0: Store initial (bad) centers
iter0_centers <- init_centers

# Iteration 1: Assign points to nearest center, then compute new centers
distances <- as.matrix(dist(rbind(sample_data, init_centers)))
distances <- distances[1:nrow(sample_data), (nrow(sample_data)+1):ncol(distances)]
iter1_clusters <- apply(distances, 1, which.min)
iter1_centers <- t(sapply(1:3, function(k) {
  if (sum(iter1_clusters == k) > 0) {
    colMeans(sample_data[iter1_clusters == k, , drop = FALSE])
  } else {
    init_centers[k, ]
  }
}))

# Iteration 2: Reassign and recompute
distances2 <- as.matrix(dist(rbind(sample_data, iter1_centers)))
distances2 <- distances2[1:nrow(sample_data), (nrow(sample_data)+1):ncol(distances2)]
iter2_clusters <- apply(distances2, 1, which.min)
iter2_centers <- t(sapply(1:3, function(k) {
  if (sum(iter2_clusters == k) > 0) {
    colMeans(sample_data[iter2_clusters == k, , drop = FALSE])
  } else {
    iter1_centers[k, ]
  }
}))

# Final: Converged solution (for comparison)
sample_data_df <- as_tibble(sample_data)
km_final <- kmeans(sample_data_df, centers = 3, nstart = 25, iter.max = 100)
final_clusters <- km_final$cluster
final_centers <- km_final$centers

# Convert to tibbles for plotting
sample_data <- as_tibble(sample_data)
iter0_centers <- as_tibble(iter0_centers)
iter1_centers <- as_tibble(iter1_centers)
iter2_centers <- as_tibble(iter2_centers)
final_centers <- as_tibble(final_centers)
colnames(iter0_centers) <- colnames(iter1_centers) <- colnames(iter2_centers) <- c("median_income", "unemployment")

# Create visualization panels for 4 stages
plot_data_iter0 <- sample_data %>%
  mutate(cluster = factor(iter1_clusters),  # Use iter1 assignments
         stage = "Iter 0: Bad Start")

plot_data_iter1 <- sample_data %>%
  mutate(cluster = factor(iter1_clusters),
         stage = "Iter 1: First Move")

plot_data_iter2 <- sample_data %>%
  mutate(cluster = factor(iter2_clusters),
         stage = "Iter 2: Improving")

plot_data_final <- sample_data %>%
  mutate(cluster = factor(final_clusters),
         stage = "Converged")

plot_data_all <- bind_rows(plot_data_iter0, plot_data_iter1, plot_data_iter2, plot_data_final) %>%
  mutate(stage = factor(stage, levels = c("Iter 0: Bad Start", 
                                          "Iter 1: First Move", 
                                          "Iter 2: Improving",
                                          "Converged")))

centers_data <- bind_rows(
  iter0_centers %>% mutate(stage = "Iter 0: Bad Start"),
  iter1_centers %>% mutate(stage = "Iter 1: First Move"),
  iter2_centers %>% mutate(stage = "Iter 2: Improving"),
  final_centers %>% mutate(stage = "Converged")
) %>%
  mutate(stage = factor(stage, levels = c("Iter 0: Bad Start",
                                          "Iter 1: First Move",
                                          "Iter 2: Improving",
                                          "Converged")))

ggplot(plot_data_all, aes(x = median_income, y = unemployment, color = cluster)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_point(data = centers_data, aes(x = median_income, y = unemployment),
             shape = 4, size = 6, stroke = 2, color = "black") +
  scale_x_continuous(labels = dollar_format()) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  scale_color_viridis_d() +
  facet_wrap(~stage, ncol = 4) +
  labs(title = "k-Means Algorithm: Iterative Convergence from Terrible Initialization",
       subtitle = "Black X marks show centroids moving from completely wrong positions (left) to optimal locations (right)",
       x = "Median Household Income",
       y = "Unemployment Rate") +
  theme_minimal() +
  theme(legend.position = "none",
        strip.text = element_text(size = 10, face = "bold"))
```

::: {.fragment}
**Notice**: Starting with all three centers clustered together (poor initialization), they progressively spread out and move toward the true cluster centers. This shows why k-means uses `nstart = 25` to try multiple random initializations!
:::

## k-Means in Action: County Clustering {.smaller}

```{r}
# Prepare data for clustering
county_features <- cluster_data %>%
  select(median_income, unemployment) %>%
  scale()  # Standardize features (critical for k-means!)

# Fit k-means with k=3
set.seed(42)
kmeans_result <- kmeans(county_features, centers = 3, nstart = 25)

# Add cluster assignments back to original data
cluster_data <- cluster_data %>%
  mutate(assigned_cluster = factor(kmeans_result$cluster))

# Examine cluster centers (in original scale)
cluster_centers <- cluster_data %>%
  group_by(assigned_cluster) %>%
  summarise(
    median_income = mean(median_income),
    unemployment = mean(unemployment),
    n_counties = n()
  )

knitr::kable(cluster_centers, 
             digits = 0,
             col.names = c("Cluster", "Avg Income", "Avg Unemployment (%)", "# Counties"),
             caption = "k-Means Cluster Profiles")
```

::: {.fragment}
**Interpretation**: k-means successfully identified three distinct county types!
:::

## Visualizing k-Means Results {.smaller}

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

# Create cluster centers in original scale
centers_df <- as_tibble(kmeans_result$centers) %>%
  mutate(
    median_income = median_income * sd(cluster_data$median_income) + mean(cluster_data$median_income),
    unemployment = unemployment * sd(cluster_data$unemployment) + mean(cluster_data$unemployment),
    assigned_cluster = factor(1:3)
  )

ggplot(cluster_data, aes(x = median_income, y = unemployment, 
                        color = assigned_cluster, shape = assigned_cluster)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_point(data = centers_df, aes(x = median_income, y = unemployment),
             size = 8, shape = 4, stroke = 2, color = "black") +
  scale_x_continuous(labels = dollar_format()) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  scale_color_viridis_d(name = "Cluster", labels = c("1: Urban", "2: Rural", "3: Suburban")) +
  scale_shape_discrete(name = "Cluster", labels = c("1: Urban", "2: Rural", "3: Suburban")) +
  labs(title = "k-Means Clustering Results: County Economic Profiles",
       subtitle = "Black X marks show cluster centers (centroids)",
       x = "Median Household Income",
       y = "Unemployment Rate") +
  theme_minimal() +
  theme(legend.position = "right")
```

::: {.fragment}
**Policy application**: Now we can design different interventions for each cluster type!
:::

## The Crucial Question: How Many Clusters (k)? {.smaller}

::: {.fragment}
**Problem**: k-means requires YOU to choose k upfront. But how do you know the "right" k?
:::

::: {.fragment}
**Bad news**: There's no single "correct" answer—depends on your goals!
:::

::: {.fragment}
**Good news**: We have methods to guide the choice:
:::

::: {.fragment}
**1. Elbow Method**: Plot within-cluster sum of squares (WCSS) vs k

- WCSS always decreases as k increases
- Look for "elbow" where adding more clusters gives diminishing returns
:::

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5

# Calculate WCSS for different k values
wcss_values <- map_dbl(1:10, function(k) {
  km <- kmeans(county_features, centers = k, nstart = 25)
  km$tot.withinss
})

elbow_data <- tibble(
  k = 1:10,
  wcss = wcss_values
)

ggplot(elbow_data, aes(x = k, y = wcss)) +
  geom_line(size = 1.2, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  geom_vline(xintercept = 3, linetype = "dashed", color = "red", alpha = 0.6) +
  annotate("text", x = 3.3, y = max(elbow_data$wcss) * 0.8, 
           label = "Elbow at k=3", color = "red", hjust = 0, size = 5) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = "Elbow Method: Choosing k",
       subtitle = "Look for the 'elbow' where improvement slows",
       x = "Number of Clusters (k)",
       y = "Within-Cluster Sum of Squares (WCSS)") +
  theme_minimal()
```

::: {.fragment}
**Here**: Clear elbow at k=3, suggesting 3 clusters is appropriate!
:::


## k-Means: Strengths and Limitations {.smaller}

::: {layout-ncol=2}
### Strengths ✓
::: {.fragment}
- **Fast**: Scales to large datasets (millions of points)
- **Simple**: Easy to understand and implement
- **Guaranteed convergence**: Always finds a solution
- **Works well** for spherical, well-separated clusters
- **Memory efficient**: Only stores centers
:::

### Limitations ✗
::: {.fragment}
- **Must choose k** upfront (no automatic selection)
- **Sensitive to initialization** (run with multiple starts)
- **Assumes spherical clusters** (equal variance)
- **Sensitive to outliers** (outliers pull centroids)
- **Requires feature scaling** (different units problematic)
- **Hard cluster assignment** (each point gets exactly one cluster)
:::
:::

::: {.fragment}
::: callout-important
**Critical preprocessing**: ALWAYS standardize features before k-means! Otherwise, variables with larger scales dominate the distance calculations.
:::
:::

## Beyond k-Means: Hierarchical Clustering {.smaller}

::: {.fragment}
**Problem with k-means**: Must choose k upfront. What if we want to explore multiple levels of grouping?
:::

::: {.fragment}
**Hierarchical clustering** builds a tree (dendrogram) showing nested clusters at all levels!
:::

::: {.fragment}
**Two approaches**:

1. **Agglomerative (bottom-up)**: Start with each point as its own cluster, merge similar ones
2. **Divisive (top-down)**: Start with all points in one cluster, split recursively
:::

::: {.fragment}
**Agglomerative algorithm** (most common):

1. Start: Each observation is its own cluster (n clusters)
2. Find the two closest clusters
3. Merge them into one cluster
4. Repeat steps 2-3 until only one cluster remains
5. Result: A tree showing all possible clusterings!
:::

::: {.fragment}
**Key choice**: How do we measure "closeness" between clusters (linkage)?
:::

## Linkage Methods {.smaller}

::: {.fragment}
**When merging clusters A and B, how do we define their distance?**
:::

::: {layout-ncol=2}
::: {.fragment}
### Single Linkage
Distance = minimum distance between any two points

$$d(A, B) = \min_{i \in A, j \in B} d(i, j)$$

**Tends to**: Create long, stringy clusters (chaining)
:::

::: {.fragment}
### Complete Linkage
Distance = maximum distance between any two points

$$d(A, B) = \max_{i \in A, j \in B} d(i, j)$$

**Tends to**: Create compact, spherical clusters
:::

::: {.fragment}
### Average Linkage
Distance = average of all pairwise distances

$$d(A, B) = \frac{1}{|A||B|}\sum_{i \in A, j \in B} d(i, j)$$

**Tends to**: Balance between single and complete
:::

::: {.fragment}
### Ward's Method
Minimize within-cluster variance (like k-means objective)

**Tends to**: Create equal-sized clusters (most popular!)
:::
:::

## Hierarchical Clustering in Action {.smaller}

```{r}
# Perform hierarchical clustering
hc_result <- hclust(dist(county_features), method = "ward.D2")

# Cut tree to get 3 clusters
hc_clusters <- cutree(hc_result, k = 3)

# Add to data
cluster_data <- cluster_data %>%
  mutate(hc_cluster = factor(hc_clusters))
```

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 6

# Plot dendrogram
fviz_dend(hc_result, k = 3, 
          cex = 0.5,
          color_labels_by_k = TRUE,
          rect = TRUE,
          rect_border = "steelblue",
          rect_fill = TRUE,
          main = "Hierarchical Clustering Dendrogram (Ward's Method)",
          xlab = "Counties",
          ylab = "Height (Distance)",
          sub = "Red boxes show the 3 clusters when tree is cut at height ≈ 6")
```

::: {.fragment}
**Reading the dendrogram**: Height shows how different clusters are. Lower merges = more similar.
:::

## Comparing k-Means vs Hierarchical {.smaller}

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 5

# Compare the two clustering methods
p1 <- ggplot(cluster_data, aes(x = median_income, y = unemployment, 
                               color = assigned_cluster)) +
  geom_point(size = 2.5, alpha = 0.7) +
  scale_x_continuous(labels = dollar_format()) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  scale_color_viridis_d(name = "Cluster") +
  labs(title = "k-Means Clustering",
       x = "Median Income", y = "Unemployment Rate") +
  theme_minimal() +
  theme(legend.position = "bottom")

p2 <- ggplot(cluster_data, aes(x = median_income, y = unemployment,
                               color = hc_cluster)) +
  geom_point(size = 2.5, alpha = 0.7) +
  scale_x_continuous(labels = dollar_format()) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  scale_color_viridis_d(name = "Cluster") +
  labs(title = "Hierarchical Clustering (Ward)",
       x = "Median Income", y = "Unemployment Rate") +
  theme_minimal() +
  theme(legend.position = "bottom")

p1 + p2
```

::: {.fragment}
**Notice**: Both methods find similar groups, but boundaries differ slightly. This is normal!
:::

## When to Use Each Method {.smaller}

```{r}
#| echo: false

comparison_table <- tibble(
  Aspect = c("Speed", "Scalability", "Choose k", "Cluster Shape", 
             "Interpretability", "Deterministic", "When to Use"),
  `k-Means` = c(
    "Fast (O(nkp))",
    "Millions of points",
    "Must specify upfront",
    "Assumes spherical",
    "Moderate (centers)",
    "No (random init)",
    "Large datasets, known k, spherical clusters"
  ),
  `Hierarchical` = c(
    "Slower (O(n² log n))",
    "Thousands of points max",
    "Can explore all k",
    "Any shape",
    "High (dendrogram)",
    "Yes (deterministic)",
    "Small datasets, explore structure, need dendrogram"
  )
)

knitr::kable(comparison_table, caption = "k-Means vs Hierarchical Clustering")
```

::: {.fragment}
::: callout-tip
**Practical advice**: Try k-means first (fast, scales well). Use hierarchical clustering for smaller datasets when you want to explore the hierarchy or create visualizations for stakeholders.
:::
:::

# Part III: Principal Component Analysis (PCA)

## The Dimensionality Problem {.smaller}

::: {.fragment}
**Scenario**: You have survey data with 50 questions measuring community well-being
:::

::: {.fragment}
**Challenges**:

- 50 dimensions are hard to visualize
- Many questions are correlated (ask similar things)
- Hard to interpret what "matters most"
- Models may overfit with so many features
:::

::: {.fragment}
**Question**: Can we reduce 50 dimensions to 5-10 **without losing much information**?
:::

::: {.fragment}
**Answer**: Yes! **Principal Component Analysis (PCA)**
:::

::: {.fragment}
::: callout-note
**Key insight**: PCA finds new variables (principal components) that capture the maximum variance in the data. The first few components often capture most of the information!
:::
:::

## What PCA Does: Geometric Intuition {.smaller}

::: {.fragment}
**PCA finds new coordinate axes that align with the directions of maximum variance**:
:::

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

# Create correlated 2D data for visualization
set.seed(789)
n <- 100
x1 <- rnorm(n)
x2 <- 0.8 * x1 + rnorm(n, 0, 0.3)

pca_demo_data <- tibble(x1, x2)
pca_result_demo <- prcomp(pca_demo_data, scale = TRUE)

# Extract PC directions
pc1_direction <- pca_result_demo$rotation[, 1]
pc2_direction <- pca_result_demo$rotation[, 2]

# Create arrows for PC directions
arrow_length <- 3
arrows <- tibble(
  x = c(0, 0),
  y = c(0, 0),
  xend = c(pc1_direction[1] * arrow_length, pc2_direction[1] * arrow_length),
  yend = c(pc1_direction[2] * arrow_length, pc2_direction[2] * arrow_length),
  PC = c("PC1 (87% variance)", "PC2 (13% variance)")
)

ggplot(pca_demo_data, aes(x = x1, y = x2)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_segment(data = arrows, aes(x = x, y = y, xend = xend, yend = yend, color = PC),
               arrow = arrow(length = unit(0.3, "inches"), type = "closed"),
               size = 1.5) +
  scale_color_manual(values = c("PC1 (87% variance)" = "red", 
                                "PC2 (13% variance)" = "blue"),
                     name = "Principal Components") +
  coord_equal() +
  labs(title = "PCA Finds Directions of Maximum Variance",
       subtitle = "PC1 (red) captures most variation, PC2 (blue) captures remaining variation",
       x = "Original Variable 1", y = "Original Variable 2") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

::: {.fragment}
**Key property**: Principal components are **orthogonal** (uncorrelated) to each other!
:::

## PCA: The Mathematics {.smaller}

::: {.fragment}
**PCA finds new variables (principal components) as linear combinations of original variables**:
:::

::: {.fragment}
$$\text{PC}_1 = w_{11}X_1 + w_{12}X_2 + \cdots + w_{1p}X_p$$
$$\text{PC}_2 = w_{21}X_1 + w_{22}X_2 + \cdots + w_{2p}X_p$$
$$\vdots$$
$$\text{PC}_p = w_{p1}X_1 + w_{p2}X_2 + \cdots + w_{pp}X_p$$
:::

::: {.fragment}
**Constraints**:

1. **PC1 has maximum variance** among all possible linear combinations
2. **PC2 has maximum variance** among all combinations orthogonal to PC1
3. **PC3 has maximum variance** among all combinations orthogonal to PC1 and PC2
4. ... and so on

Each component is a **weighted average** of the original features!
:::

::: {.fragment}
**Technical note**: The weights ($w_{ij}$) come from the eigenvectors of the covariance matrix. PCA is fundamentally an eigenvalue decomposition!
:::

## PCA on Real Policy Data {.smaller}

::: {.fragment}
**Example**: County-level data with **20 socioeconomic variables**

This is where PCA shines - too many dimensions to visualize or interpret!
:::

```{r}
# Create high-dimensional county dataset (20 variables!)
set.seed(999)
n_counties_pca <- 150

county_data <- tibble(
  county_id = 1:n_counties_pca,
  # Economic indicators
  median_income = rnorm(n_counties_pca, 55000, 15000),
  unemployment_rate = rnorm(n_counties_pca, 5.5, 2),
  poverty_rate = rnorm(n_counties_pca, 12, 4),
  gini_index = rnorm(n_counties_pca, 0.45, 0.05),
  median_home_value = rnorm(n_counties_pca, 250000, 80000),
  
  # Demographics
  median_age = rnorm(n_counties_pca, 38, 8),
  percent_white = rnorm(n_counties_pca, 65, 20),
  percent_hispanic = rnorm(n_counties_pca, 18, 12),
  percent_black = rnorm(n_counties_pca, 13, 10),
  population_density = rexp(n_counties_pca, rate = 1/500),
  
  # Education
  percent_college = rnorm(n_counties_pca, 35, 10),
  percent_hs_grad = rnorm(n_counties_pca, 88, 6),
  per_pupil_spending = rnorm(n_counties_pca, 12000, 3000),
  
  # Health
  life_expectancy = rnorm(n_counties_pca, 78, 3),
  percent_insured = rnorm(n_counties_pca, 90, 5),
  obesity_rate = rnorm(n_counties_pca, 30, 6),
  
  # Infrastructure
  broadband_access = rnorm(n_counties_pca, 82, 12),
  commute_time_mins = rnorm(n_counties_pca, 25, 8),
  
  # Political
  voter_turnout = rnorm(n_counties_pca, 65, 10),
  dem_vote_share = rnorm(n_counties_pca, 48, 15)
) %>%
  mutate(
    # Create realistic correlations
    unemployment_rate = pmax(1, 15 - 0.0001 * median_income + unemployment_rate),
    percent_college = pmax(10, 10 + 0.0003 * median_income + percent_college),
    poverty_rate = pmax(0, 30 - 0.0002 * median_income + poverty_rate),
    median_home_value = pmax(50000, median_home_value + 2 * median_income),
    life_expectancy = pmax(70, life_expectancy + 0.00005 * median_income),
    broadband_access = pmax(40, pmin(98, broadband_access + 0.0002 * population_density)),
    obesity_rate = pmax(15, obesity_rate - 0.0001 * median_income)
  )

# Show structure
county_data %>% 
  select(1:6) %>%
  head(3)
```

::: {.fragment}
**Challenge**: How do we understand patterns in 20-dimensional space? 🤯
:::

## Performing PCA {.smaller}

```{r}
# Standardize and run PCA
county_features_pca <- county_data %>%
  select(-county_id) %>%
  scale()

pca_result <- prcomp(county_features_pca)

# Variance explained by each component
variance_explained <- tibble(
  PC = paste0("PC", 1:length(pca_result$sdev)),
  variance = pca_result$sdev^2,
  prop_variance = variance / sum(variance),
  cumulative_variance = cumsum(prop_variance)
)

# Show first 8 components (enough to see the pattern)
knitr::kable(head(variance_explained, 8), 
             digits = 3,
             col.names = c("Component", "Variance", "Prop. Variance", "Cumulative"),
             caption = "Variance Explained by Principal Components (first 8 of 20)")
```

::: {.fragment}
**Key finding**: First 5 components capture ~60% of variance. First 10 capture ~85%!

We've reduced 20 dimensions to 5 while retaining most information.
:::

## Visualizing Variance Explained {.smaller}

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 5

# Scree plot (first 10 components for clarity)
variance_plot <- variance_explained %>%
  head(10) %>%
  mutate(PC = factor(PC, levels = PC))

p1 <- ggplot(variance_plot, aes(x = PC, y = prop_variance)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_line(aes(group = 1), color = "red", size = 1) +
  geom_point(color = "red", size = 3) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "Scree Plot: Variance by Component",
       subtitle = "Clear 'elbow' after PC1 suggests keeping 1 component",
       x = "Principal Component",
       y = "Proportion of Variance Explained") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0))

# Cumulative variance
p2 <- ggplot(variance_plot, aes(x = PC, y = cumulative_variance)) +
  geom_col(fill = "darkgreen", alpha = 0.8) +
  geom_line(aes(group = 1), color = "red", size = 1) +
  geom_point(color = "red", size = 3) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "blue", alpha = 0.6) +
  annotate("text", x = 7, y = 0.82, label = "80% threshold", color = "blue", size = 4) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  labs(title = "Cumulative Variance Explained",
       subtitle = "First 6 PCs capture ~50%",
       x = "Principal Component",
       y = "Cumulative Proportion") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0))

p1 + p2
```

::: {.fragment}
**Decision**: We have to pick the first N components that capture enough variance, how far do we go?
:::

## Interpreting Principal Components: Loadings {.smaller}

::: {.fragment}
**Question**: What do the principal components *mean*? 
:::

::: {.fragment}
**Answer**: Look at the **loadings** (weights) on each original variable!
:::

```{r}
# Extract and display loadings for first 3 PCs
loadings <- as_tibble(pca_result$rotation[, 1:3], rownames = "Variable") %>%
  rename(PC1 = PC1, PC2 = PC2, PC3 = PC3) %>%
  arrange(desc(abs(PC1)))

knitr::kable(head(loadings, 12), 
             digits = 3,
             caption = "Top 12 PCA Loadings: How Original Variables Map to Components")
```


## Visualizing Loadings (Biplot) {.smaller}

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7

# Create biplot with selected key variables (avoid clutter with 20 variables)
key_vars <- c("median_income", "unemployment_rate", "poverty_rate", 
              "percent_college", "population_density", "life_expectancy",
              "median_home_value", "broadband_access")

# Get loadings for key variables
key_loadings <- pca_result$rotation[key_vars, 1:2] * 4  # Scale arrows for visibility

# PC scores for counties
pc_scores <- as_tibble(pca_result$x[, 1:2])

ggplot(pc_scores, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.4, size = 2, color = "gray50") +
  # Add loading arrows
  geom_segment(data = as_tibble(key_loadings, rownames = "Variable"),
               aes(x = 0, y = 0, xend = PC1, xend = PC2),
               arrow = arrow(length = unit(0.3, "cm")),
               color = "red", size = 1, alpha = 0.7) +
  geom_text(data = as_tibble(key_loadings, rownames = "Variable"),
            aes(x = PC1, y = PC2, label = Variable),
            color = "darkred", size = 3.5, fontface = "bold",
            nudge_x = 0.5, nudge_y = 0.3) +
  labs(title = "PCA Biplot: Counties in PC1-PC2 Space",
       subtitle = "Red arrows show key variables driving the principal components (8 of 20 shown)",
       x = paste0("PC1 (", round(variance_explained$prop_variance[1] * 100, 1), "% variance)"),
       y = paste0("PC2 (", round(variance_explained$prop_variance[2] * 100, 1), "% variance)")) +
  theme_minimal()
```

::: {.fragment}
**Reading the biplot**: 

- Each point is a county in the reduced 2D space
- Arrows show original variables; longer arrows = stronger contribution to PCs
- Variables pointing in same direction are correlated
:::

## Using PCA for Visualization {.smaller}

::: {.fragment}
**One of PCA's best uses**: Visualize high-dimensional data in 2D or 3D!
:::

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

# Add PC scores back to original data
county_data_with_pcs <- county_data %>%
  bind_cols(as_tibble(pca_result$x[, 1:3])) %>%
  mutate(
    income_category = case_when(
      median_income < 45000 ~ "Low Income",
      median_income < 65000 ~ "Middle Income",
      TRUE ~ "High Income"
    ),
    income_category = factor(income_category, 
                            levels = c("Low Income", "Middle Income", "High Income"))
  )

ggplot(county_data_with_pcs, aes(x = PC1, y = PC2, color = income_category, size = population_density)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c("Low Income" = "coral", 
                                "Middle Income" = "steelblue",
                                "High Income" = "darkgreen"),
                     name = "Income Category") +
  scale_size_continuous(name = "Population\nDensity", range = c(1, 6)) +
  labs(title = "Counties Visualized in Principal Component Space",
       subtitle = "20 dimensions reduced to 2 for easy visualization",
       x = paste0("PC1: Economic Prosperity (", round(variance_explained$prop_variance[1] * 100, 1), "%)"),
       y = paste0("PC2: Urbanization (", round(variance_explained$prop_variance[2] * 100, 1), "%)")) +
  theme_minimal() +
  theme(legend.position = "right")
```

::: {.fragment}
**Insight**: Clear separation by income in PC1, with urban/rural distinction in PC2!
:::

## PCA: Strengths and Limitations {.smaller}

::: {layout-ncol=2}
### Strengths ✓
::: {.fragment}
- **Reduces dimensionality** while preserving variance
- **Reveals correlation structure** in data
- **Enables visualization** of high-dimensional data
- **No labels needed** (unsupervised)
- **Removes multicollinearity** (PCs are uncorrelated)
- **Data compression** (store fewer numbers)
- **Feature engineering** (use PCs as features in models)
:::

### Limitations ✗
::: {.fragment}
- **Assumes linearity** (linear combinations only)
- **Sensitive to scale** (must standardize!)
- **Hard to interpret** (what is "PC3"?)
- **Loses original feature meaning**
- **Assumes Gaussian structure** works best
- **Not robust to outliers**
- **No sparsity** (all variables contribute to all PCs)
:::
:::

::: {.fragment}
::: callout-important
**Critical preprocessing**: ALWAYS standardize features before PCA! Variables with larger scales will dominate the principal components.
:::
:::

## When to Use PCA in Policy Work {.smaller}

::: {.fragment}
**PCA is useful for**:
:::

::: {layout-ncol=2}
::: {.fragment}
### Good Use Cases
**1. Survey dimension reduction**
- 100 survey questions → 10 underlying themes
- Example: Well-being survey

**2. Visualization**
- Plot high-dimensional data in 2D
- Example: County comparison dashboard

**3. Noise reduction**
- Remove low-variance components
- Example: Sensor data preprocessing

**4. Multicollinearity**
- Create uncorrelated features for regression
- Example: Economic indicators
:::

::: {.fragment}
### When NOT to Use
**1. Need interpretability**
- PCs are hard to explain to stakeholders
- Use domain-specific indices instead

**2. Categorical variables**
- PCA assumes continuous variables
- Use Multiple Correspondence Analysis (MCA)

**3. Non-linear structure**
- PCA only finds linear combinations
- Try kernel PCA or autoencoders for non-linear reduction

**4. Sparse data (many zeros)**
- Standard PCA doesn't handle well
- Use sparse PCA or matrix factorization
:::
:::

## Alternative Dimensionality Reduction Methods {.smaller}

::: {.fragment}
**PCA isn't the only option**:
:::

::: {.fragment}
```{r}
#| echo: false

dim_red_methods <- tibble(
  Method = c("PCA", "Factor Analysis", "ICA", "Multidimensional Scaling", "t-SNE", "UMAP"),
  `Best For` = c(
    "Linear structure, data compression, feature engineering",
    "Identifying latent factors (psychology/surveys)",
    "Independent sources (signal processing)",
    "Preserving pairwise distances (e.g., city proximity)",
    "2D/3D visualization only (non-linear)",
    "2D/3D visualization only (faster than t-SNE)"
  ),
  Speed = c("Fast", "Medium", "Medium", "Slow", "Slow", "Fast"),
  `Can Use for Modeling` = c("Yes", "Yes", "Yes", "Yes", "No (viz only)", "No (viz only)"),
  Notes = c(
    "Maximizes variance",
    "Latent variables",
    "Statistical independence",
    "Multidimensional Scaling",
    "Stochastic, non-linear",
    "Better global structure"
  )
)

knitr::kable(dim_red_methods, caption = "Dimensionality Reduction vs Visualization Methods")
```
:::

::: {.fragment}
::: callout-important
**Critical distinction**: t-SNE and UMAP are **visualization-only** techniques that reduce to 2D/3D for plotting. They're stochastic, non-invertible, and don't preserve distances globally—you can't use their outputs as features in downstream models. PCA, Factor Analysis, and ICA produce general-purpose reduced dimensions suitable for modeling.
:::
:::

::: {.fragment}
::: callout-tip
**For policy work**: Use **PCA** when you need interpretable dimensions for modeling. Use **t-SNE/UMAP** only for creating exploratory 2D visualizations of complex data.
:::
:::

# Part IV: Combining Clustering and PCA

## The Power of Combination {.smaller}

::: {.fragment}
**Common workflow**: Use PCA THEN cluster on the principal components!
:::

::: {.fragment}
**Why this works well**:

1. **PCA removes noise** → cleaner clustering
2. **PCA removes correlation** → better distance metrics
3. **Fewer dimensions** → faster clustering
4. **Visualize in 2D** → see clusters clearly
:::

```{r}
# Cluster on first 3 principal components
set.seed(42)
pcs_for_clustering <- pca_result$x[, 1:3]
kmeans_on_pcs <- kmeans(pcs_for_clustering, centers = 4, nstart = 25)

# Add cluster assignments
county_data_with_pcs <- county_data_with_pcs %>%
  mutate(cluster = factor(kmeans_on_pcs$cluster))
```

## Visualizing Clusters in PC Space {.smaller}

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

ggplot(county_data_with_pcs, aes(x = PC1, y = PC2, color = cluster, shape = cluster)) +
  geom_point(size = 4, alpha = 0.8) +
  stat_ellipse(aes(group = cluster), type = "norm", level = 0.7, size = 1) +
  scale_color_viridis_d(name = "Cluster") +
  scale_shape_manual(values = c(16, 17, 15, 18), name = "Cluster") +
  labs(title = "k-Means Clustering on Principal Components",
       subtitle = "Clustering in reduced PC space often works better than in original space",
       x = "PC1: Economic Prosperity",
       y = "PC2: Urbanization") +
  theme_minimal() +
  theme(legend.position = "right")
```

::: {.fragment}
**Result**: Clear, interpretable clusters in the 2D PC space!
:::

# Part V: Practical Guidelines and Best Practices

## Complete Unsupervised Learning Workflow {.smaller}

::: {.fragment}
**Recommended step-by-step process**:
:::

::: {.fragment}
**1. Exploratory Data Analysis**
```r
# Check distributions, correlations, outliers
summary(data)
cor(data) %>% corrplot()
```
:::

::: {.fragment}
**2. Preprocessing**
```r
# Handle missing data, remove outliers, standardize
data_clean <- data %>%
  drop_na() %>%
  mutate(across(where(is.numeric), scale))
```
:::

::: {.fragment}
**3. Dimensionality Assessment**
```r
# If p > 10-20, consider PCA first
pca_result <- prcomp(data_clean)
# Keep components explaining 80-90% variance
```
:::

::: {.fragment}
**4. Choose Clustering Method**
```r
# Try multiple k values, compare metrics
fviz_nbclust(data_clean, kmeans, method = "silhouette")
```
:::

## Validation and Interpretation {.smaller}

::: {.fragment}
**5. Cluster Validation**
:::

::: {.fragment}
- **Internal metrics**: Silhouette, within-cluster SS, Dunn index
- **Stability**: Do clusters persist across bootstrap samples?
- **Separability**: Are clusters well-separated?
:::

::: {.fragment}
**6. Domain Validation**
:::

::: {.fragment}
- **Profile clusters**: What characterizes each group?
- **Check face validity**: Do clusters make sense to domain experts?
- **Compare to known groupings**: Do clusters align with existing taxonomies?
:::

::: {.fragment}
**7. Communication**
:::

::: {.fragment}
- **Visualize in 2D**: Use PCA or t-SNE for stakeholder-friendly plots
- **Name clusters**: Give meaningful labels ("Urban Thriving," not "Cluster 3")
- **Provide profiles**: Show average characteristics of each cluster
- **Show examples**: Highlight representative members of each cluster
:::

## Common Pitfalls to Avoid {.smaller}

::: {layout-ncol=2}
### Technical Mistakes
::: {.fragment}
1. **Not standardizing** features
   - Different scales distort distances
   
2. **Using raw data with outliers**
   - Outliers pull centroids, distort PCs
   
3. **Ignoring correlations**
   - Correlated features dominate distance
   
4. **Choosing k arbitrarily**
   - Use elbow, silhouette, domain knowledge
   
5. **Running k-means once**
   - Use `nstart = 25` for multiple initializations
:::

### Interpretation Mistakes
::: {.fragment}
6. **Over-interpreting PCs**
   - PC3 with 5% variance is mostly noise
   
7. **Treating clusters as "truth"**
   - Clusters are useful abstractions, not natural law
   
8. **Ignoring within-cluster variation**
   - Not all members perfectly fit their cluster
   
9. **Forgetting context**
   - Numbers don't replace domain expertise
   
10. **Not validating results**
    - Always check if clusters make sense
:::
:::

## Real-World Policy Application: Putting It All Together {.smaller}

::: {.fragment}
**Case Study**: State education policy segmentation
:::

::: {.fragment}
**Goal**: Group US states by education system characteristics to design targeted federal programs
:::

::: {.fragment}
**Data**: 50 states × 20 variables (spending, outcomes, demographics, policies)
:::

::: {.fragment}
**Workflow**:

1. **PCA**: Reduce 20 variables to 5 PCs (85% variance)
   - PC1: "Resource level" (spending, teacher salaries)
   - PC2: "Student demographics" (poverty, diversity)
   - PC3: "Performance" (test scores, graduation)
   
2. **Clustering**: k-means on PCs, k=5 clusters
   - Use silhouette to validate k=5
   
3. **Profiling**: Characterize each cluster
   - Cluster 1: "High-resource, high-performing"
   - Cluster 2: "Low-resource, diverse student body"
   - etc.
   
4. **Policy design**: Tailor interventions to each cluster type
:::

## Ethical Considerations in Unsupervised Learning {.smaller}

::: {.fragment}
**Unsupervised learning isn't neutral**!
:::

::: {.fragment}
**Potential harms**:

1. **Reinforcing stereotypes**
   - Clusters may reflect historical discrimination
   
2. **Unequal treatment**
   - Different interventions for different clusters
   - Ensure equitable access to resources
   
3. **Loss of individuality**
   - Treating everyone in a cluster the same
   - Allow opt-out or appeals
   
4. **Opaque decision-making**
   - Hard to explain why someone is in Cluster 3
   - Document methodology, allow inspection
:::

::: {.fragment}
::: callout-important
**Policy principle**: Clustering should inform, not determine, decisions about individuals. Always combine algorithmic groupings with human judgment!
:::
:::

## Advanced Topics: What's Next {.smaller}

::: {.fragment}
**Beyond the basics**:
:::

::: {.fragment}
**Trend: Deep learning for unsupervised learning**

- Variational autoencoders (VAEs)
- Contrastive learning (SimCLR)
- Self-supervised learning
- Foundation models (pre-training without labels)
:::

## Looking Ahead to Final Week {.smaller}

::: {.fragment}
**Week 10 (next class - Dec 1)**: AI-Augmented Policy Analysis
:::

::: {.fragment}
- Integrating LLM APIs into data workflows
- Using AI assistants for code generation and debugging
- Prompt engineering for data analysis tasks
- Ethical use of AI tools in research
- Modern development environments (Cursor, GitHub Copilot)
:::

::: {.fragment}
**Final Exam (Dec 3)**: In-class, cumulative, difficult.

:::

::: {.fragment}
**Assignment 3 Due (Dec 6 at 8:00 AM)**: Final project report
- Labor Action Tracker predictions
- Model selection and evaluation
- Policy recommendations
- ≤4 pages/1500 words
:::

## Key Takeaways {.smaller}

::: {.fragment}
1. **Unsupervised learning discovers patterns without labels**—essential for exploration and segmentation
:::

::: {.fragment}
2. **k-means clustering** is fast and scalable but requires choosing k and assumes spherical clusters
:::

::: {.fragment}
3. **Hierarchical clustering** shows nested structure via dendrograms but doesn't scale to large datasets
:::

::: {.fragment}
4. **PCA reduces dimensionality** by finding directions of maximum variance—great for visualization and noise reduction
:::

::: {.fragment}
5. **Combining PCA + clustering** often works better than clustering in original high-dimensional space
:::

::: {.fragment}
6. **Always standardize** features before clustering or PCA!
:::

::: {.fragment}
7. **Validation is subjective**—use internal metrics + domain knowledge + stakeholder input
:::

::: {.fragment}
8. **Clusters are abstractions**, not truth—use them to inform decisions, not make them automatically
:::

## Questions?

::: {.fragment}
**Remember**: Unsupervised learning is a powerful exploratory tool. It helps us understand structure in complex policy data, but the insights must always be validated with domain expertise and used responsibly in decision-making!
:::

::: {.fragment}
**Next class**: We'll explore how modern AI tools (LLMs, GitHub Copilot, Cursor) can augment your data science workflow—and how to use them ethically and effectively!
:::

---

